\chapter{Introduction}

\section{Prerequisites}

This text assumes familiarity with metric space theory and with linear algebra. 

\section{Linear algebra notation}

\subsection{Some important vector spaces}

\begin{definition}
	We write $\R^n$ to denote the vector space of lists of $n$ numbers. We will sometimes write its elements as horizontal lists with commas, as in $(a_1, \dotsc, a_n)$, and sometimes as vertical column vectors, as in \[ \begin{bmatrix} a_1 \\ \vdots \\ a_n \end{bmatrix}.  \]
\end{definition}

\begin{definition}
	We write $\mathscr{M}_{n,m}$ for the set of all $n \times m$ matrices (ie, the matrices with $n$ rows and $m$ columns). This is naturally a vector space. 
\end{definition}

\begin{definition}
	If $V$ and $W$ are both vector spaces, we write $\mathscr{L}(V, W)$ for the set of all linear maps $V \to W$. This is naturally a vector space. 
\end{definition}

\subsection{Matrix notation}

\begin{definition}
	Suppose $V$ is a finite dimensional vector space and $B = ( v_1, \dotsc, v_n )$ is a basis for $V$. Then any $v \in V$ can be written uniquely as $a_1v_1 + \dotsb + a_n v_n$ for some scalars $a_1, \dotsc, a_n \in \R$, and we define the \emph{representation of $v$ with respect to $B$}, denoted $[v]_B$, to be the column vector with $n$ entries that records the scalars $a_1, \dotsc, a_n$. In other words,
\[ [v]_B = \begin{bmatrix} a_1 \\ \vdots \\ a_n \end{bmatrix}. \]
\end{definition}

\begin{lemma}
	If $V$ is an $n$-dimensional vector space and $B = (v_1, \dotsc, v_n)$ is a basis for $V$, then the function $V \to \R^n$ given by $v \mapsto [v]_B$ is an isomorphism, with inverse given by
	\[ \begin{bmatrix} a_1 \\ \vdots \\ a_n \end{bmatrix} \mapsto a_1 v_1 + \dotsb + a_n v_n. \]
\end{lemma}

\begin{definition}
Suppose $\ell : V \to W$ is a linear map between finite dimensional vector spaces. Let $B = v_1, \dotsc, v_m$ be a basis for $V$ and let $C$ be a basis for $W$. The \emph{matrix representation of $\ell$ with respect to $B$ and $C$}, denoted $[\ell]_{B,C}$, is the $n \times m$ matrix whose $i$th column of $[\ell]_{B,C}$ is $[\ell(v_i)]_C$. In other words,
\[ [\ell]_{B,C} = \begin{bmatrix} [\ell(v_1)]_C & \dotsb & [\ell(v_m)]_C \end{bmatrix}. \]
\end{definition}

\begin{lemma}
	If $\ell : V \to W$ is a linear map between finite dimensional vector spaces and $B$ and $C$ are bases for $V$ and $W$, respectively, then \[ [\ell]_{B,C}[v]_B = [\ell(v)]_C \]
	for any $v  \in V$. 
\end{lemma}

\begin{lemma}
	If $\ell : U \to V$ and $\ell' :  V \to W$ are linear maps between finite dimensional vector spaces, and $A, B$ and $C$ are bases for $U, V$, and $W$, respectively, then 
	\[ [\ell' \circ \ell]_{A,C} = [\ell']_{B,C}[\ell]_{A,B}. \]
\end{lemma}

\begin{lemma}
	If $\ell : V \to W$ is a linear map between finite dimensional vector spaces and $B$ and $C$ are bases for $V$ and $W$, respectively, then the rank of the matrix $[\ell]_{B,C}$ equals the dimension of the range $\ell(V)$.
\end{lemma}

\begin{lemma}
	Suppose that $V$ and $W$ are $m$- and $n$-dimensional vector spaces, respectively, and that and $B$ and $C$ are bases for $V$ and $W$, respectively. Then the function $\ell \mapsto [\ell]_{B,C}$ is an isomorphism \[ \begin{tikzcd} \mathscr{L}(V, W) \ar{r} & \mathscr{M}_{n,m}. \end{tikzcd} \] 
\end{lemma}

\begin{definition}
	If $\ell : \R^m \to \R^n$ is a linear map, the \emph{standard matrix representation} of $\ell$, denoted $[\ell]$, is the matrix of $\ell$ with respect to the standard bases on $\R^m$ and $\R^n$.
\end{definition} 

\begin{lemma}
	If $A$ is a $n \times m$ matrix and $\ell_A : \R^m \to \R^n$ is the linear map $\ell_A(v) = Av$, then \[ [\ell_A] = A. \]
\end{lemma} 

\section{Little-oh notation}

Throughout this section, let $X$ be a metric space and let $x_0$ be a point in $X$. 

\begin{definition}
	A function $g : X \setminus \{x_0\} \to \R$ is \emph{positive} if $g(x) > 0$ for all $x \in X \setminus \{x_0\}$. Similarly, $g$ is \emph{non-negative} if $g(x) geq 0$ for all $x \in X \setminus \{x_0\}$.
\end{definition}

Suppose $g$ is a positive function. \Cref{little-o-definition} below formulates what it means for a non-negative function $f : X \setminus \{x_0\} \to \R$ to be ``little-oh of $g$ as $x$ tends to $x_0$.'' Intuitively, this condition means that $f$ is \emph{much smaller} than $g$ near $x_0$. 

\begin{definition} \label{little-o-definition}
	A function $f : X \setminus \{x_0\} \to \R$ is \emph{little-oh of $g$ as $x$ tends to $x_0$}, written \[ f(x) = o(g(x)) \text{ as } x \to x_0 \] 
	if, for every $\epsilon > 0$, there exists an open neighborhood $U$ of $x_0$ such that $f(x) \leq \epsilon g(x)$ for all $x \in U \setminus \{x_0\}$. Equivalently, this means that 
	\[ \lim_{x \to x_0} \frac{|f(x)|}{g(x)} = 0. \]
	When $x_0$ can be inferred from context, we write simply $f = o(g)$. 
\end{definition} 

\begin{remark}
	It's worth noting that the use of the symbol ``$=$'' above is mathematically abusive. The left-hand side of the ``$=$'' is a function and the right-hand side is a property of functions; of course, a function cannot literally be equal to a property. Rather, the ``$=$'' is being used here to mean that the function on the left-hand side has the property described on the right-hand side. As annoying as it is, this abuse of notation is fairly standard, so it's probably best to just get used to it. 
\end{remark}

\begin{exercise} \label{little-o-equivalent-exercise}
	Prove that the two conditions in \cref{little-o-definition} are in fact equivalent. 
\end{exercise}

\begin{exercise}
	Let $X = \R$ and $x_0 = 0$. For each of the functions $f$ and $g$ described below, sketch graphs of $f$ and $g$ and then determine whether or not $f = o(g)$ as $x \to 0$.
	\begin{enumerate}[(a)]
		\item $f(x) = x$ and $g(x) = x^2$. 
		\item $f(x) = x$ and $g(x) = |x|$. 
		\item $f(x) = x^2$ and $g(x) = |x|$. 
	\end{enumerate}
\end{exercise}

\begin{exercise}
	Let $X = \R^2$ and $x_0 = 0$. For each of the following functions $f$, determine whether or not $f(x) = o(|x|)$ as $x \to 0$. 
	\begin{enumerate}[(a)]
		\item $f(x,y) = x - y$. 
		\item $f(x,y) = x^2 + y^2$. 
	\end{enumerate}
\end{exercise}

\begin{exercise}[``Smaller than small is still small''] \label{less-than-small-is-small}
	Suppose $f_1, f_2 : X \setminus \{x_0\} \to \R$ are functions such that $|f_1(x)| \leq |f_2(x)|$ for all $x$ in a punctured neighborhood of $x_0$, and $f_2(x) = o(g(x))$ as $x \to x_0$. Then $f_1(x) = o(g(x))$ also. 
\end{exercise}

\begin{exercise} \label{little-o-vector-space}
	For a fixed positive function $g : X \setminus \{x_0\} \to \R$, prove that the set of functions $f : X \setminus \{x_0\} \to \R$ which are $o(g)$ as $x \to x_0$ is a vector space under the natural operations. In other words, verify the following three facts. 
	\begin{enumerate}[(1)]
		\item The zero function is $o(g(x))$. 
		\item (``Scalar multiples of small are still small'') If $c \in \R$ is a scalar and $f(x) = o(g(x))$, then 
		\[ cf(x) = o(g(x)). \] 
		\item (``Sum of smalls is still small'') If $f_1(x) = o(g(x))$ and $f_2(x) = o(g(x))$, then \[ (f_1 + f_2)(x) = o(g(x)). \]
	\end{enumerate}
\end{exercise}

\begin{exercise} \label{little-o-multiplicative}
	Suppose $f_1, f_2, g_1, g_2 : X \setminus \{x_0\} \to \R$ are functions with $g_1, g_2$ positive such that $f_1(x) = o(g_1(x))$ and $f_2(x) = o(g_2(x))$ as $x \to x_0$. Prove that $(f_1 \cdot f_2)(x) = o((g_1 \cdot g_2)(x))$. 
\end{exercise}

\begin{comment}
\begin{exercise}
	Suppose $f : X \to \R$ is continuous at $x_0$ and $f(x_0) = 0$, and that $f = o(g)$ as $x \to  x_0$ for some function $g : X \setminus\{x_0\} \to \R$. Suppose further that $s : \R \to \R$ is a function such that $s = o(|r|)$ as $r \to 0$. Then $s \circ f = o(g)$ as $x \to x_0$. 
\end{exercise}

\begin{proof}
	Suppose $\epsilon > 0$. We want to find an open neighborhood $U$ of $x_0$ such that
	\[ |s(f(x))| \leq \epsilon|t(g(x))| \]
	for $x \in U$. 
	
	Since $s$ is $o(t)$ as $r \to 0$, we know that for any $\epsilon_1 > 0$ there exists $\delta_1 > 0$ such that $|r| < \delta_1$ implies that
	\[ \begin{aligned} |s(r)| \leq \epsilon_1|t(r)|. \end{aligned} \] 
	Since $f(x_0) = 0$ and $f$ is continuous at $x_0$, there exists a neighborhood $U_1$ of $x_0$ such that
	\[ |f(x)| < \delta_1  \]
	for all $x \in U_1$. This means that for $x \in U_1$, we have
	\[ |s(f(x))| \leq \epsilon_1|t(f(x))|. \]
	Since $f = o(g)$ as $x \to x_0$, for any $\epsilon_2 > 0$ there is another neighborhood $U_2$ of $x_0$ such that $|f(x)| \leq \epsilon_2 |g(x)|$. This means that
	\[ |s(f(x))| \leq \epsilon_1 \epsilon_2 |g(x)|. \]
	
	there then exists $\delta_2 > 0$ such that $|h| < \delta$ implies that $|df_a(h) + r(h)| < \delta_1$. Then for $|h| < \delta_2$, we have
	\[ |s(df_a(h) + r(h)| \leq \epsilon_1 |df_a(h) + r(h)| 
	\leq \epsilon_1|f'(a)||h| + \epsilon_1|r(h)| \]
	Since $r$ is $o(|h|)$, for any $\epsilon_2 > 0$ there exists $\delta_3 > 0$ such that $|h| < \delta_3$ implies that $|r(h)| \leq \epsilon_2 |h|$. Let $\delta = \min\{\delta_2, \delta_3\}$. Then for $|h| < \delta$, we can continue the above chain of inequalities to obtain the following.  
	\[ |s(df_a(h) + r(h)| \leq \epsilon_1|f'(a)||h| + \epsilon_1 \epsilon_2|h| = \epsilon_1(|f'(a)| + \epsilon_2)|h| \]
	So we let $\epsilon_2 = 1$ and then let 
	\[ \epsilon_1 = \frac{\epsilon}{|f'(a)| + 1}. \]
	The $\delta$ obtained by going through the above process then has the desired property.
\end{proof}
\end{comment}

% TODO: exercise about composition? i'm not sure how to formuulate the correct statement though

\section{Norms of linear maps}

\begin{definition}
	Suppose $\ell : \R^m \to \R^n$ is a linear map. Then we define the \emph{operator norm}, or just \emph{norm}, of $\ell$, denoted $\|\ell\|$, by
	\[ \|\ell\| = \sup\, \{ |\ell(v)| : |v| \leq 1 \}. \]
\end{definition} 

\begin{exercise}
	Suppose $\ell : \R^m \to \R^n$ is a linear map. Show that 
	\[ \|\ell\| = \inf \, \{ \lambda \in \R : |\ell(v)| \leq \lambda |v| \text{ for all } v \in \R^m \}. \]
	In particular, $|\ell(v)| \leq \|\ell\| \cdot |v|$ for all $v \in \R^m$. 
\end{exercise}

\begin{exercise} \label{norm-is-hausdorff}
	Suppose $\ell : \R^m \to \R^n$ is a linear map. Show that $\ell = 0$ if and only if $\|\ell\| = 0$. 
\end{exercise}

\begin{exercise}[Subadditivity] \label{subadditive}
	Suppose $\ell, \ell' :  \R^m \to \R^n$ are both linear maps. Show that
	\[ \|\ell + \ell'\| \leq \|\ell\| + \|\ell'\|. \]
\end{exercise}

\begin{definition} \label{operators-as-metric-space}
	Let $\mathscr{L}(\R^m, \R^n)$ denote the set of all linear maps $\R^m \to \R^n$. Define the distance between $\ell, \ell' \in \mathscr{L}(\R^m, \R^n)$ by $\|\ell - \ell'\|$. It follows from \cref{norm-is-hausdorff,subadditive} that this distance function makes $\mathscr{L}(\R^m, \R^n)$ a metric space. 
\end{definition}

\begin{exercise}[``Evaluation is continuous''] \label{evaluation-continuous}
	Suppose $v_0 \in \R^m$ is a fixed vector. Then the ``evaluate at $v_0$'' function $\mathscr{L}(\R^m, \R^n) \to \R^n$ given by $\phi \mapsto \phi(v_0)$ is continuous. \textit{Possible hint}. Since the ``evaluate at $v_0$'' function is linear, it is sufficient to check continuity at 0. 
\end{exercise}

\begin{exercise}[Submultiplicativity] \label{norm-is-submultiplicative}
	Suppose $\ell : \R^m \to \R^n$ and $\ell' : \R^n \to \R^p$ are both linear maps. Show that \[ \|\ell' \circ \ell\| \leq \|\ell'\| \cdot \|\ell\|. \]
	Give an example to show that this inequality can be strict. 
\end{exercise}

\begin{exercise}[``Composition is continuous''] \label{composition-continuous}
	\begin{enumerate}[(a)]
		\item Suppose $\ell : \R^n \to \R^p$ is a linear map. Show that the ``postcompose with $\ell$'' function $\mathscr{L}(\R^m, \R^n) \to \mathscr{L}(\R^m, \R^p)$ given by $\phi \mapsto \ell \circ \phi$ is continuous. 
		\item Suppose $\ell : \R^m \to \R^n$ is a linear map. Show that the ``precompose with $\ell$'' function $\mathscr{L}(\R^n, \R^p) \to \mathscr{L}(\R^m, \R^p)$ given by $\phi \mapsto \phi \circ \ell$ is continuous. 
	\end{enumerate}
	\noindent \textit{Possible hint} (for both parts). Because both the ``postcompose with $\ell$'' and the ``precompose with $\ell$'' maps are linear, it is sufficient to show that the maps are continuous at 0. You will likely find \cref{norm-is-submultiplicative} useful. 
\end{exercise}

\begin{exercise}[``Linear functions of small functions are still small''] \label{linear-of-small-is-small}
	Suppose $r : \R^m \to \R^n$ is a function such that $|r(v)| = o(|v|)$ as $v \to 0$. If $\ell : \R^n \to \R^p$ is linear, prove that $|\ell(r(v))| = o(|v|)$ also. 
\end{exercise}

\begin{comment}
\begin{proof}
	Observe that
	\[ \frac{|\ell(r(v))|}{|v|} \leq \frac{\|\ell\| \cdot |r(v)|}{|v|}. \]
	Since $|r(v)| = o(|v|)$, we see that the right-hand side tends to 0 as $v \to 0$. The squeeze theorem thus guarantees that $|\ell(r(v))| = o(|v|)$ also. 
\end{proof}
\end{comment}

\begin{lemma} \label{norm-is-realized}
	Suppose $\ell : \R^m \to \R^n$ is a nonzero linear map. Then there exists $v_0 \in \R^m$ such that $|v_0| = 1$ and $|\ell(v_0)| = \|\ell\|$. 
\end{lemma}

\begin{proof}
	The ``unit disk'' \[ D = \{ x \in \R^m : |x| \leq 1 \} \]
	is a nonempty, connected, and compact subset of $\R^m$. The function $v \mapsto |\ell(v)|$ is also a continuous function $D \to \R$, so its image $|\ell(D)|$ must also be a nonempty, connected, and compact subset of $\R$. In other words, it must be a closed interval in $\R$. This means that there exists $v_0 \in D$ such that 
	\[ |\ell(v_0)| = \sup |\ell(D)| = \|\ell\|. \]
	Since $\ell \neq 0$, we know $\|\ell\| \neq 0$ from \cref{norm-is-hausdorff}, which in turn means that $v_0 \neq 0$. To prove that $|v_0| = 1$, assume for a contradiction that $|v_0| < 1$, and let $c = 1/|v_0| > 1$. Then $|cv_0| = |c||v_0| = 1$, so $cv_0 \in D$. But then
	\[ |\ell(cv_0)| = |c\ell(v_0)| = c|\ell(v_0)| > |\ell(v_0)| = \sup |\ell(D)|, \]
	which is absurd. 
\end{proof}

\begin{exercise} \label{linear-and-small}
	Suppose $\ell : \R^m \to \R^n$ is a linear map and $|\ell(v)| = o(|v|)$ as $v \to 0$. Prove that $\ell = 0$. 
\end{exercise}

\begin{comment}
	The fact that $|\ell(v)| = o(|v|)$ tells us that
	\begin{equation} \label{big-sequence} \lim_{v \to 0} \frac{|\ell(v)|}{|v|} = 0, \end{equation}
	using \cref{little-o-equivalent}. Suppose for a contradiction that $\ell \neq 0$. Using \cref{norm-is-realized}, choose $v_0 \in \R^m$ such that $|v_0| = 1$ and $|\ell(v_0)| = \|\ell\|$. For scalars $c$, observe that $cv_0 \to 0$ as $c \to 0$, which means that \cref{big-sequence} implies that
	\[ \lim_{c \to 0} \frac{|\ell(cv_0)|}{|cv_0|} = 0.  \]
	But observe that
	\[ \frac{|\ell(cv_0)|}{|cv_0|} = \frac{|c\ell(v_0)|}{|cv_0|} = \|\ell\| \]
	so
	\[ 0 = \lim_{c \to 0} \frac{|\ell(cv_0)|}{|cv_0|} = \lim_{c \to 0} \|\ell\| = \|\ell\|. \]
	This contradicts \cref{norm-is-hausdorff}. 
\end{comment}