\chapter{Manifolds}

Manifolds allow us to formalize the idea of geometry without having to make a choice of coordinates. The idea, roughly speaking, is to have a space and then remember \emph{all possible} choice of coordinates on that space! 

At this point, we will need to use the language of topological spaces. If you have seen metric spaces but not topological spaces, you should look over \cref{topological-spaces-basics} briefly (in particular, at least \cref{topological-space-definition,metric-to-topological,continuous-definition,open-map-definition,homeomorphism-definition}).

\section{Definition of a manifold}

Throughout, let $X$ be a topological space.\index{topological space}

\begin{definition}[Chart] \index{chart}
	A \emph{chart} on $X$ is a pair $(U, x)$ consisting of an open subset $U \subseteq X$ and a continuous, open\index{open!open map}, and injective function $x : U \to \R^n$ for some $n$. We call the integer $n$ the \emph{dimension} of $(U,x)$, and write $\dim (U,x)$ to denote it. If $p \in X$, we say that the chart $(U,x)$ \emph{contains} $p$ if $p \in U$. Also, we call the functions $x_i := \pi_i \circ x : U \to \R$ \emph{coordinate functions} on $U$ with respect to the chart $(U, x)$, and for any $p \in U$, the list of numbers $x_1(p), \dotsc, x_n(p)$ are called the \emph{coordinates} of $p$ with respect to the chart $(U, x)$.\index{coordinates} 
\end{definition}

\begin{exercise} \label{polar-representation}
	Let $X = \R^2$ and let $U$ denote $\R^2$ minus the non-negative real axis. Let $\text{polar} : U \to \R^2$ be the function which sends a point $p$ to its polar representation $(r(p), \theta(p))$, where $r(p) = |p|$ and $\theta(p)$ is the angle in radians strictly between 0 and $2\pi$ formed between $P$ and the non-negative real axis. Show that $(U, \text{polar})$ is a chart. 
\end{exercise}

\begin{definition}[Transition function] \label{transition-map} \index{transition function}
	Suppose $(U, x)$ and $(U, x')$ are charts on $X$. The \emph{transition function} from $(U, x)$ to $(U, x')$ is the function $x' \circ x^{-1}$. 
	\[ \begin{tikzcd}
	x(U \cap V) \ar{r}{x^{-1}} & U \cap V \ar{r}{x'} & x'(U \cap V)
	\end{tikzcd} \]
\end{definition}

\begin{definition}[Compatibility of charts] \index{compatibile} \label{compatibility}
	The two charts $(U, x)$ and $(U', x')$ on $X$ are \emph{smoothly compatible}, or just \emph{compatible}, if both of the transition maps between $(U, x)$ and $(U', x')$ are smooth. 
\end{definition}

\begin{exercise} \label{polar-compatible-with-identity}
	Let $\id : \R^2 \to \R^2$ denote the identity map and $\text{polar} : U \to \R^2$ the polar representation map from \cref{polar-representation}. Show that $(\R^2, \id)$ and $(U, \text{polar})$ are compatible charts. 
\end{exercise}

\begin{exercise} \label{compatibility-not-equivalence-relation}
	Show that the relation ``is compatible with'' is \emph{not} an equivalence relation on the set of all charts. 
	\begin{hint}
		The function $\R \to \R$ given by $x \mapsto x^3$ is a smooth homeomorphism, but its inverse is not differentiable. 
	\end{hint}
\end{exercise}

\begin{exercise} \label{dimension-well-defined}
	Suppose $(U, x)$ and $(U', x')$ are compatible charts and that $U \cap U'$ is nonempty. Show that $\dim(U,x) = \dim(U',x')$. 
	\begin{hint}
		Pick a point $p \in U \cap V$, and consider the derivatives of $x' \circ x^{-1}$ at $x(p)$ and of $x \circ x'^{-1}$ at $x'(p)$. 
	\end{hint}
\end{exercise}

\begin{definition}[Atlas] \index{atlas} \index{atlas!maximal atlas} \index{maximal atlas|see {atlas, maximal atlas}}
	An \emph{atlas} on $X$ is a collection $\mathscr{A}$ of charts such that covers $X$. In other words, \[ \bigcup_{(U, x) \in \mathscr{A}} U = X. \]
	An atlas is \emph{maximal} if there is no strictly larger atlas containing it.
\end{definition}

\begin{exercise}[Existence and uniqueness of maximal atlases] \label{existence-maximal-atlas}
	Prove that every atlas $\mathscr{A}$ is contained in a unique maximal atlas.\index{atlas!maximal atlas} 
	\begin{hint}
		Let $\mathscr{A}'$ be the set of all charts that are compatible with all of the charts in $\mathscr{A}$. Prove that $\mathscr{A}'$ is an atlas (caution: keep \cref{compatibility-not-equivalence-relation} in mind). Then prove that it is maximal, and that it is the only maximal atlas containing $\mathscr{A}$. 
	\end{hint}
\end{exercise}

This leads us to the following definition.

\begin{definition}[Manifold] \index{manifold} \index{smooth manifold|see {manifold}}
	A \emph{smooth manifold}, or just a \emph{manifold}, is a pair $(X, \mathscr{A}_X)$ where $X$ is a topological space and $\mathscr{A}_X$ is a maximal atlas. Often, write simply ``$X$'' in place of the pair $(X, \mathscr{A}_X)$, and sometimes $\mathscr{A}_X$ is called the ``manifold structure on $X$.''
\end{definition}

Of course, \cref{existence-maximal-atlas} tells us that any not-necessarily-maximal atlas on a metric space $X$ defines a manifold structure on $X$ in a canonical way. When constructing examples, we will usually give an explicit non-maximal atlas and then use \cref{existence-maximal-atlas} to extend it to a maximal atlas. But we require maximality in the definition in part because, when dealing with abstract manifolds in proofs, we will often find it convenient to assume that the atlas that the manifold comes equipped with is already maximal.

We can also make the following definition.

\begin{definition}[Dimension] \index{dimension!of a manifold}
	Suppose $X$ is a manifold and $p \in X$ is a point. By \cref{dimension-well-defined}, there is a unique integer $n$ such that $\dim(U,x) = n$ for every chart $(U, x) \in \mathscr{A}_X$ containing $p$. This integer is called the \emph{dimension of $X$ at $p$}, and is denoted $\dim_p(X)$. We say that $X$ is \emph{$n$-dimensional} if $\dim_p(X) = n$ for all $p \in X$.
\end{definition}

\begin{unimportantremark}
	For any $k = 0, 1, 2, \dotsc, \infty$, we could define charts to be \emph{$C^k$-compatible} by requiring that the transition functions between them are $C^k$ (rather than smooth, as in \cref{compatibility}). This would lead to a definition of \emph{$C^k$-manifolds}. In these notes, we'll just stick to the $k = \infty$ case. This is by far the most interesting.  
\end{unimportantremark}

\section{Examples}

\subsection{First examples}

\begin{example} \label{natural-euclidean-atlas}
	The single chart $(\R^n, \id)$, where $\id : \R^n \to \R^n$ is the identity map, defines an atlas on $\R^n$. We can extend this to a maximal atlas, called the \emph{euclidean atlas} on $\R^n$. Unless explicitly specified otherwise, we always regard $\R^n$ as a manifold by equipping it with the euclidean atlas. By \cref{polar-compatible-with-identity}, the polar representation chart $(U, \text{polar})$ is an element of the euclidean atlas. 
\end{example}

\begin{exercise}
	Let $f : \R \to \R$ be the function $f(x) = x^3$. Note that $f$ is smooth and strictly increasing (hence injective and open, by \cref{strictly-monotone-iff-injective,strictly-monotone-implies-open}). Show that $(\R, f)$ is \emph{not} a member of the euclidean atlas on $\R$. 
\end{exercise}

\begin{exercise} \label{euclidean-chart-characterization}
	Suppose $U$ is an open subset of $\R^n$ and $f : U \to \R^n$ is a function. Show that the pair $(U, f)$ is a chart in the euclidean atlas on $\R^n$ if and only if $f$ is smooth, injective, and \'etale.
\end{exercise}

\begin{exercise}[The circle $S^1$] \index{s1@$S^1$|see {circle}} \index{circle} \label{circle-projection-atlas}
	Let $S^1$ denote the set of points in $\R^2$ along the unit circle. In other words,
	\[ S^1 = \{ (a, b) \in \R^2 : a^2 + b^2 = 1 \}. \]
	\begin{enumerate}[(a)]
		\item Do \cref{metric-subspace}. This shows that both ways that we might think to regard $S^1$ as a topological space are actually the same. 
		
		\item Let $U = S^1 \setminus \{(0,1)\}$. For $p \in S^1$, let $x(p)$ denote the $x$-coordinate of the point where the line $y = -1$ intersects the line passing through $(0,1)$ and $p$. See \cref{circle-projection}. If $p = (a, b)$, find a formula for $x(p)$ in terms of $a$ and $b$. 
		
		\begin{figure}
			\begin{center}
				\begin{tikzpicture}
				\begin{axis}[
				xmin=-1.5,xmax=1.5,
				ymin=-1.3,ymax=1.3,
				ytick={-1,1},
				xtick={-1,1},
				xticklabels={},
				yticklabels={},
				axis lines=middle,
				axis line style=lightgray,
				width=0.75\textwidth,
				axis equal
				]
				
				\addplot[black,style=thick,-] expression[domain=-1:1,samples=100]{sqrt(1-x^2)}; 
				\addplot[black,style=thick,-] expression[domain=-1:1,samples=100]{-sqrt(1-x^2)};
				
				
				\addplot[red,<->] expression[domain=-0.1:0.95,samples=2]{-2.414*x+1}; 
				
				
				\addplot[black,<->] expression[domain=-1.5:1.5,samples=2]{-1};
				
				\node[label={85:{$(0,1)$}},color=white,draw=black,circle,fill,inner sep=2pt] at (axis cs:0,1) {}; 
				\node[label={0:{$p$}},color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:0.707,-0.707) {}; 
				\node[color=red,draw=red,circle,fill,inner sep=2pt] at (axis cs:0.8285,-1) {}; 
				
				\end{axis}
				\end{tikzpicture}
			\end{center}
			\caption{Given a point $p \in U = S^1 \setminus \{(0,1)\}$, we draw a line connecting $p$ to $(0,1)$, indicated in red. The point where that line intersects the line $y = -1$ (in black) is sometimes called the \emph{projection of $p$ from $(0,1)$ onto $\R \times \{-1\}$}. The $x$-coordinate of this point is $x(p)$.}  \label{circle-projection}
		\end{figure}
		
		\item Show that $x : U \to \R$ is a homeomorphism.\index{homeomorphism} Thus $(U, x)$ is a chart on $S^1$. 
		\begin{hint} 
			Use geometry to write down a formula for the inverse function to $x$. 
		\end{hint}
		
		\item Let $U' = S^1 \setminus \{(0,-1)\}$. For $p \in U'$, let $x'(p)$ denote the $x$-coordinate of the point where the line $y = 1$ intersects the line passing through $(0,-1)$ and $p$. Show that $x' : U' \to \R$ is also a homeomorphism. 
		
		\item Show that $(U, x)$ and $(U', x')$ are compatible charts. Thus $\mathscr{A} = \{(U, x), (U', x')\}$ is an atlas on $S^1$. Extending $\mathscr{A}$ to a maximal atlas $\mathscr{A}_{S^1}$, we have defined a manifold structure on $S^1$.
		
		\item Let $V = \{ (a,b) \in S^1 : a > 0 \}$ be the ``right half of the circle.'' If $y : V \to \R$ is the function $y(a,b) = b$, show that $(V,y)$ is a chart and that it is compatible with the charts in the atlas $\mathscr{A}$ above. Thus $(V,y) \in \mathscr{A}_{S^1}$. 
		
		\item Can you think of any other charts in the maximal atlas $\mathscr{A}_{S^1}$, besides the three discussed above? 
	\end{enumerate}
\end{exercise}

\begin{exercise}[The $n$-sphere $S^n$] \label{sphere-projection-atlas} \index{sn@$S^n$|see {$n$-sphere}} \index{n-sphere@$n$-sphere}
	Let $S^n$ denote the set of points in $\R^{n+1}$ of norm 1. Generalize \cref{circle-projection-atlas} and show that ``projection from the north pole'' and ``projection from the south pole'' define an atlas on $S^n$. 
\end{exercise}

\begin{example}[Matrices]
	There is an isomorphism of vector spaces $x : \mathscr{M}_{n \times m} \to \R^{mn}$\index{mnm@$\mathscr{M}_{n \times m}$} where, if $A \in \mathscr{M}_{n \times m}$ has entries $a_{i,j}$, then
	\[ x(A) = \begin{bmatrix} a_{1,1} \\ a_{1,2} \\ \vdots \\ a_{1,m} \\ a_{2,1} \\ \vdots \\ a_{n,m} \end{bmatrix}. \]
	If we endow $\mathscr{M}_{n \times m}$ with the canonical topology,\index{canonical topology} then $x$ is automatically a homeomorphism by \cref{linear-isomorphism-homeomorphism}.\index{homeomorphism} In other words, the single chart $(\mathscr{M}_{n \times m}, x)$ is an atlas on $\mathscr{M}_{n \times m}$. Extending it to a maximal atlas using \cref{existence-maximal-atlas}, we see that we can regard $\mathscr{M}_{n,m}$ as a manifold. It is $mn$-dimensional. The same process allows us to regard \emph{any} finite dimensional vector space (not just $\mathscr{M}_{n \times m}$) as a manifold.  
\end{example}

\begin{exercise}[Graphs] \label{graphs-continuous}
	Suppose $U$ is an open subset of $\R^m$ and $f : U \to \R^n$ is a continuous function. 
	\begin{enumerate}[(a)]
		\item There are two ways we might think to regard $U \times \R^n$ as a topological space. One way is to use the product topology \ref{product-space}. The second way is to regard $U \times \R^n$ as a subset of $\R^m \times \R^n = \R^{m+n}$ and then give $U \times \R^n$ the subspace topology \ref{subspace-topology}. Show that these two topologies are the same. 
		
		\item Let
		\[ \Gamma = \{ (x,y) \in U \times \R^n : f(x) = y \} \]
		be the graph of $f$, regarded as a subspace of $U \times \R^n$. Show that the function $\pi : \Gamma \to \R^m$ given by $\pi(x,y) = x$ is continuous, injective, and open. 
	\end{enumerate}
	The single chart $(\Gamma, \pi)$ defines an atlas $\Gamma$. Extending this to a maximal atlas, we can regard $\Gamma$ as a manifold. 
\end{exercise}

\begin{exercise}[Open submanifolds] \label{open-submanifold}
	Let $X$ be a manifold and $U \subseteq X$ an open subset. Show that \[ \mathscr{A}_U = \{ (V,x) \in \mathscr{A}_X : V \subseteq U \} \]
	is a maximal atlas on $U$. We call $U$, equipped with this maximal atlas, an \emph{open submanifold} of $X$. 
\end{exercise}

\begin{exercise}[Product manifold]
	Let $X$ and $Y$ be manifolds. Let $\mathscr{A}$ be the set of charts of the form $(U \times V, x \times y)$, where $(U,x : U \to \R^m) \in \mathscr{A}_X$, $(V, y : V \to \R^n) \in \mathscr{A}_Y$, and $x \times y$ denotes the natural map $U \times V \to \R^m \times \R^n = \R^{m+n}$. Show that $\mathscr{A}$ is an atlas on $X \times Y$. We call $X \times Y$, equipped with the corresponding maximal atlas, the \emph{product manifold}. 
\end{exercise}

\subsection{M\"obius strip}

Let $B$ denote the box $[0,1] \times (0,1)$ inside $\R^2$, and let $\sim$ be the equivalence relation on $B$ where $(a,b) \sim (a',b')$ if we either have $a = a'$ and $b = b'$, or else we have $a = 0$, $a' = 1$, and $b' = 1-b$. See \cref{mobius-box}. We will denote the equivalence class of a point $(a, b) \in B$ as $[a, b]$. 

\begin{figure}
	\begin{center}
	\begin{tikzpicture}
	\begin{axis}[
	xmin=-0.1,xmax=1.1,
	ymin=-0.1,ymax=1.1,
	ytick={1},
	xtick={1},
	xticklabels={1},
	yticklabels={1},
	axis lines=middle,
	axis line style=lightgray,
	axis equal,
	width=0.6\textwidth
	]
	
	\addplot[black, opacity=0.2, domain=0:1, samples=2] {0.25};
	\addplot[black, opacity=0.2, domain=0:1, samples=2] {0.75};
	
	\foreach \n in {1, ..., 9}
		\addplot +[mark=none, black, solid, opacity=0.2] coordinates {(0.1*\n, 0) (0.1*\n, 1)};
	
	\addplot[black, dashed, name path=top, domain=0:1, samples=2] {0};
	\addplot[black, dashed, name path=bottom, domain=0:1, samples=2] {1};
	\addplot [fill=red, fill opacity=0.1] 
	fill between[of=top and bottom];
	
	
	\addplot[red, thick, domain=0:1, samples=2, opacity=0.5] {0.5};
	\addplot +[mark=none, solid, thick, black] coordinates {(0, 0) (0, 1)};
	\addplot +[mark=none, solid, thick, black] coordinates {(1, 0) (1, 1)};
	
	\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:0,0.25) {}; 
	\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:1,0.75) {}; 
	\end{axis}
	\end{tikzpicture}
	
	\vspace{5em}
	
	
	\caption{Above is a picture of the box $B = [0,1] \times (0,1)$ and the equivalence relation $\sim$ described above. Geometrically, this equivalence relation identifies the vertical black line on the left of the box with the vertical black line on the right of the box, but ``with a twist.'' For example, the point $(0,1/4)$ on the left is identified with the point $(1,3/4)$ on the right. If we pick up the box and actually glue together the edges as indicated by the equivalence relation, we end up with the M\"obius strip as pictured below.} \label{mobius-box}
	
	\vspace{5em}
	
	\begin{tikzpicture}
	\begin{axis}[
	hide axis,
	view={40}{40},
	width=\textwidth]
	
	\addplot3 [surf,
	red,
	opacity=0.1,
	faceted color=black,
	point meta=x,
	samples=40,
	samples y=5,
	z buffer=sort,
	domain=0:360,
	y domain=-0.5:0.5
	] (
	{(1+0.5*y*cos(x/2)))*cos(x)},
	{(1+0.5*y*cos(x/2)))*sin(x)},
	{0.5*y*sin(x/2)});
	
	% Center line
	\addplot3 [samples=50, domain=-145:180, samples y=0, red, thick, opacity=0.5
	] ({cos(x)},{sin(x)},{0});
	
	% Outside line
	\addplot3 [samples=50, domain=-135:495, samples y=0, black, dashed] (
	{(1+0.25*cos(x/2)))*cos(x)},
	{(1+0.25*cos(x/2)))*sin(x)},
	{0.25*sin(x/2)});
	
	% Glued line
	\addplot3 [samples=50,domain=-0.5:0.5,samples y=0, black, thick,
	] ({1+0.5*x},{0},{0});
	
	% Point
	\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:1+1/8,0,0) {};
	\end{axis}
	\end{tikzpicture}
	\end{center}
\end{figure}

\begin{exercise}
	Check that $\sim$ is an equivalence relation, and that every equivalence class contains either one or two points. 
\end{exercise}

The \emph{M\"obius strip}\index{mobius strip@M\"obius strip} $M$ is the set of equivalence classes in $B$. In other words, we take the box $[0,1] \times (0,1)$ and ``glue'' its left and right edges, with a twist. If you've ever made a M\"obius strip out of paper, you can probably see that we've formalized exactly that same process. See \cref{mobius-box}, again.

We regard $M$ as a topological space using the quotient topology, as in \cref{quotient-space}. To give $X$ the structure of a manifold, we will construct an atlas consisting of two charts. But, before proceeding with the formalism, let's describe the intuition. Imagine making a cut in the M\"obius strip corresponding to a vertical cut in the box $B$. Once we make such a cut, we can untwist the M\"obius strip and we're left with an open rectangle, which we can regard as an open subset of $\R^2$. In other words, we've defined a chart on $M$. The points along the cut are missing from this chart; but if we construct a second chart by making a second separate cut, we will have constructed two charts that cover $M$. 

Let's formalize this now. The easiest place to cut is the line along which we glued the M\"obius strip in the first place. In other words, we let \[ U = \{ [a, b] \in M : a \in (0,1) \} \] 
and let $x : U \to \R^2$ be the function $x([a,b]) = (a, b)$.

\begin{exercise}
	Show that $U$ is an open subset of $M$, and that the function $x : U \to \R^2$ is well-defined, continuous, injective, and open. 
\end{exercise}

The second chart is a bit harder to describe mathematically, but the intuitive idea is the same. We'll cut the M\"obius strip along the line corresponding the vertical line $a = 1/2$ in $B$. Formally, let \[ V = \{ [a, b] \in M : a \neq 1/2 \} \] and let $y : V \to  \R^2$ be the function defined by 
\[ y([a,b]) = \begin{cases} (a, b) & \text{if } a < 1/2 \\ (a-1, 1-b) & \text{if } a > 1/2 \end{cases} \]

\begin{exercise}
	Show that $V$ is also an open subset of $M$, and that $y : V \to \R^2$ is well-defined, continuous, injective, and open. 
\end{exercise}

\begin{exercise}
	Show that $(U, x)$ and $(V, y)$ are compatible.
\end{exercise}

Thus $\mathscr{A} = \{(U, x), (V, y)\}$ is an atlas on the M\"obius strip $M$. Letting $\mathscr{A}_M$ denote the corresponding maximal atlas, we obtain the structure of a manifold on $M$.

\begin{exercise}
	Fix a constant $c \in [0,1)$, and let $U_c$ be the subset of $M$ where we've cut out the vertical line $a = c$. Show that $U_c$ is an open subset of $M$. Write down a formula for a well-defined function $x_c : U_c \to \R^2$ which is continuous, injective, and open; the image of $x_c$ should be an open box in $\R^2$, and you should recover the function $x$ above when $c = 0$, and the function $y$ when $c = 1/2$. Then show that $(U_c, x_c) \in \mathscr{A}_M$ for all $c \in [0,1)$. 
\end{exercise}

\subsection{More quotient examples}

Here is another example of a manifold constructed as a quotient space that you probably recognize (though maybe under the name ``donut'' rather than ``torus.'')

\begin{exercise}[Torus] \label{torus} \index{torus}
	Let $B$ denote the box $[0,1] \times [0,1]$ inside $\R^2$ and let $\sim$ denote the equivalence relation generated by $(a,0) \sim (a,1)$ and $(0,b) \sim (1,b)$ for all $a, b \in [0,1]$. The \emph{torus} $T$ is defined to be the corresponding quotient space. Describe an atlas on $T$. 
\end{exercise}

% TODO: Torus picture

But sometimes quotient spaces can be nearly impossible to visualize directly; the only way to have geometric intuition in these cases is to remember the space we started with before quotienting, and the equivalence relation we defined on it. 

\begin{exercise}[Projective line] \label{rp1} \index{projective line} \index{p1@$\bbP^1$|see {projective line}}
	Let $S^1$ denote the circle as in \cref{circle-projection-atlas}. Define an equivalence relation $\sim$ on $S^1$ which identifies antipodal points; in other words, every point on $S^1$ is declared to be equivalent to the point that is diametrically opposite it. The \emph{projective line} $\bbP^1$ is defined to be the corresponding quotient space. Describe an atlas on $\bbP^1$. 
\end{exercise}

\begin{exercise}[Projective plane] \label{rp2} \index{projective plane} \index{p2@$\bbP^2$|see {projective plane}}
	Let $S^2$ denote the 2-sphere as in \cref{sphere-projection-atlas}. Define an equivalence relation $\sim$ on $S^2$ which identifies antipodal points; in other words, every point on $S^2$ is declared to be equivalent to the point that is diametrically opposite it. The \emph{projective plane} $\bbP^2$ is defined to be the corresponding quotient space. Describe an atlas on $\bbP^2$. 
\end{exercise}

\subsection{Submanifolds}

\begin{definition}
	Suppose $S$ is a subset of a manifold $X$. Then $S$ is a \emph{submanifold} of $X$ if there exist charts $(U, x) \in \mathscr{A}_X$ covering $S$ such that 
	\[ S \cap U = \{ p \in U :  x_1(p) = x_2(p) = \dotsb = x_k(p) = 0 \} \]
	for some non-negative integer $k \leq \dim(U,x)$. We say that the chart $(U,x)$ is \emph{adapted} to $S$, and the integer $k$ is called the \emph{codimension of $S$ in $(U,x)$}. 
\end{definition}

We'll justify this terminology in a moment by showing that submanifolds are actually manifolds in their own right (cf. \cref{submanifolds-are-manifolds}). But first, let's discuss some examples. 

\begin{example}
	Consider the set $S = \{ (0,y) \in \R^2 : -1 < x < 1 \}$ inside $\R^2$. In other words, $S$ is the open interval $(-1,1)$ along the $y$-axis inside $\R^2$. Then $S$ is a submanifold. Indeed, let $U$ be the open box $(-1,1) \times (-1,1)$ inside $\R^2$, and $x : U \to \R^2$ the inclusion map. Then $(U, x)$ is a chart on $\R^2$, and 
	\[ S \cap U = \{ p \in S : x_1(p) = 0 \} \]
	so $(U,x)$ is adapted to $S$ and $S$ has codimension 1 in $(U,x)$. 
\end{example}

\begin{exercise}
	Show that each of the following is a submanifold of $\R^2$. 
	\begin{enumerate}[(a)] 
		\item The $y$-axis. 
		\item The open interval $(-1,1)$ along the $x$-axis.
		\item The line $y = x$. 
	\end{enumerate}
\end{exercise}

\begin{exercise} \label{open-submanifolds-are-submanifolds}
	Suppose $X$ is a manifold and $U$ is an open subset. Show that $U$ is a submanifold of $X$. 
\end{exercise}

\begin{proposition} \label{submanifolds-are-manifolds}
	Let $S$ be a submanifold of a manifold $X$. Suppose $(U, x) \in \mathscr{A}_X$ is adapted to $S$ and $S$ has codimension $k$ in $(U, x)$. Let $x|_S : S \cap U \to \R^{n-k}$ be the function defined by 
	\[ x|_S(p) = (x_{k+1}(p), \dotsc, x_n(p)). \]
	Regarding $S$ as a topological space using the subspace topology, the pair $(S \cap U, x|_S)$ is a chart on $S$, and the set of all charts of this form is an atlas on $S$. Thus $S$ becomes a manifold if we equip it with the corresponding maximal atlas $\mathscr{A}_S$. Finally, if $(U, x) \in \mathscr{A}_X$ is a chart adapted to $S$ such that $S$ has codimension $k$ in $S$, then
	\[ \dim_p(S) = \dim_p(X) - k \]
	for all $p \in S \cap U$. 
\end{proposition}

\begin{proof}
	The proof that $x|_S$ is continuous and injective is left as an exercise (cf. \cref{submanifolds-are-manifolds-details}). Let us show that $x|_S$ is open. Suppose $V$ is an open subset of $U$. 
	
	{\color{blue} Incomplete.}
\end{proof}

% TODO: Submanifolds are manifolds

\begin{exercise} \label{submanifolds-are-manifolds-details}
	Show that $x|_S$ is continuous and injective. 
\end{exercise}

\begin{exercise}
	Suppose $X$ is a manifold and $U$ is an open subset. On the one hand, we can equip $U$ with the maximal atlas defined in \cref{open-submanifold}. On the other hand, we can regard $U$ as a submanifold (cf. \cref{open-submanifolds-are-submanifolds}) and then use \cref{submanifolds-are-manifolds} to equip $U$ with a maximal atlas. Show that these two maximal atlases are the same. 
\end{exercise}

\subsection{Submanifolds of euclidean space}

We now have two results which let us produce lots of examples of submanifolds of euclidean space out of smooth functions (in the sense of \cref{smooth-multi}). 

First up, we have the following important theorem which goes by many names; we will call it the ``regular level set theorem,'' but some other names include the ``implicit function theorem,'' the ``preimage theorem,'' and the ``submersion theorem.'' We will later generalize this theorem. 

\begin{theorem}[Regular level set theorem] \label{regular-level-set-theorem-local}
	Suppose $U$ is an open subset of $\R^m$ and $f : U \to \R^n$ is a smooth function where $m \geq n$. If $q \in \R^n$ is a regular value of $f$, then $S = f^{-1}(q)$ is a submanifold of $U$ of dimension $m-n$. 
\end{theorem}

\begin{proof}
	By replacing $f$ with the function $x \mapsto f(x)-q$, we can assume without loss of generality that $q = 0$. Fix $p \in S$. We want to find a chart adapted to $S$ containing $p$. Since $0$ is a regular value and $m \geq n$, we know that $f$ is submersive at $p$. This means that the $n \times m$ matrix $f'(p)$ has a non-vanishing $n \times n$ minor, ie, there exist $n$ distinct integers $i_1, \dotsc, i_n$ between $1$ and $m$ such that 
	\begin{equation} \label{derivative-non-vanishing-minor} 
	\det \begin{bmatrix} \partial_{i_1} f_1(p) & \dotsb & \partial_{i_n} f_1(p) \\ \vdots & \ddots & \vdots \\ \partial_{i_1} f_n(p) & \dotsb & \partial_{i_n} f_n(p) \end{bmatrix} \neq 0. 
	\end{equation}
	By permuting the coordinates, we can assume without loss of generality that $i_1 = 1, i_2 = 2, \dotsc, i_n = n$ (cf. \cref{permutation-submanifold}). Consider the function $y : U \to \R^m$ such that, if $x = (x_1, \dotsc, x_m) \in U$, then
	\[ y(x) = (f_1(x), \dotsc, f_n(x), x_{n+1}, \dotsc, x_m). \]
	Then
	\[ y'(x) = 
	\begin{bmatrix} 
	\partial_1 f_1(x) & \dotsb & \partial_n f_1(x) & \partial_{n+1} f_1(x) & \dotsb & \partial_m f_1(x) \\ 
	\vdots & \dotsb & \vdots & \vdots & \ddots & \vdots \\ 
	\partial_1 f_n(x) & \dotsb & \partial_n f_n(x) & \partial_{n+1} f_n(x) & \dotsb & \partial_m f_n(x) \\
	0 & \dotsb & 0 & 1 & \dotsb & 0 \\ 
	\vdots & \ddots & \vdots & \vdots & \ddots & \vdots \\ 
	0 & \dotsb & 0 & 0 & \dotsb & 1
	\end{bmatrix} \]
	In other words, the $(m-n) \times m$ submatrix on the bottom left of $g'(x)$ is all zeroes, and the $(m-n) \times (m-n)$ submatrix on the bottom right is the identity matrix. This means that $\det y'(x)$ is equal, up to sign, to the determinant of the $n \times n$ in the top left, which we know from \cref{derivative-non-vanishing-minor} is nonzero when $x = p$. Thus $y$ is \'etale at $p$. 
	
	Since $f$ is smooth, so is $y$, so there exists an open neighborhood $V$ of $p$ such that $y|_V$ is \'etale. By the inverse function theorem \ref{inverse-function-theorem}, we can replace $V$ with a smaller open neighborhood of $p$ in order to further assume that $y|_V$ is injective. Then \cref{euclidean-chart-characterization} tells us that $(V, y)$ is a chart in the euclidean atlas containing $p$. Moreover, 
	\[ S \cap V = \{ p' \in V : f(p') = 0 \} = \{ p' \in V : y_1(p') = \dotsb = y_n(p') = 0 \} \]
	because $y_j = f_j$ for all $j = 1, \dotsc, n$. This shows that $(V,y)$ is adapted to $S$, and that $S$ has codimension $n$ at $p$ in $\R^m$. 
\end{proof}

\begin{exercise} \label{permutation-submanifold}
	Suppose $\sigma$ is a permutation of the set $\{1, \dotsc, m\}$. Let $f_\sigma : \R^m \to \R^m$ be the function which ``permutes the coordinates of elements of $\R^m$ according to $\sigma$,'' ie, 
	\[ f_\sigma(x_1, \dotsc, x_m) = (x_{\sigma(1)}, \dotsc, x_{\sigma(m)}). \]
	Show that a subset $S$ is a submanifold of $\R^m$ if and only if $f_\sigma(S)$ is a submanifold of $\R^m$. 
\end{exercise}

\begin{proposition} \label{graphs-smooth}
	Suppose $U$ is an open subset of $\R^m$ and $f : U \to \R^n$ is a smooth function. Then the graph of $f$ is a submanifold of $\R^{m+n}$. 
\end{proposition}

\begin{proof}
	We derive this from the regular level set theorem \ref{regular-level-set-theorem-local}. Consider the function $g : U \times \R^n \to \R^n$ given by $g(x, y) = f(x) - y$. Then the graph of $f$ is equal to $g^{-1}(0)$, so the result will follow from \cref{regular-level-set-theorem-local} provided we can show that 0 is a regular value of $g$. 
	
	In fact, $g$ is submersive at all points, so all of its values are regular! To see this, we calculate $dg$. Note that if $x = (x_1, \dotsc, x_m) \in U$ and $y = (y_1, \dotsc, y_n) \in \R^n$, then
	\[ g(x,y) = g(x_1, \dotsc, x_m, y_1, \dotsc, y_n) = (f_1(x_1,\dotsc, x_m)-y_1, \dotsc, f_n(x_1, \dotsc, x_m) - y_n) \]
	by definition of $g$, which means that
	\[ \frac{\partial g_k}{\partial x_i}(x,y) = \frac{\partial f_k}{\partial x_j}(x) = \partial_j f_k(x) \quad \text{and} \quad \frac{\partial g_k}{\partial y_j}(x,y) = -\delta_{j,k} = \begin{cases} -1 & \text{if } j = k \\ 0 & \text{if } j \neq k \end{cases} \]
	where $i = 1, \dotsc, m$ and $j, k = 1, \dotsc n$. In other words, $g'(x,y)$ is the following $n \times (m+n)$ matrix. 
	\[ g'(x,y) = 
	\begin{bmatrix} 
	\partial_1 f_1(x) & \partial_2 f_1(x) & \dotsb & \partial_m f_1(x) & -1 & 0 & \dotsb & 0 \\ 
	\partial_1 f_2(x) & \partial_2 f_2(x) & \dotsb & \partial_m f_2(x) & 0 & -1 & \dotsb & 0 \\
	\vdots & \vdots & \ddots & \vdots & \vdots & \vdots & \ddots & \vdots \\ 
	\partial_1 f_n(x) & \partial_2 f_n(x) & \dotsb & \partial_m f_n(x) & 0 & 0 & \dotsb & -1 \end{bmatrix} = \begin{bmatrix} f'(x) & -I \end{bmatrix} \]
	where $I$ denotes the $n \times n$ identity matrix. Since $I$ has rank $n$, so does $g'(x,y)$, proving that $g$ is submersive everywhere. 
\end{proof}

An important cautionary remark is that sometimes, a subspace that is a manifold can fail to be a submanifold. For example, you might notice that we showed in \cref{graphs-continuous} that the graph of a \emph{continuous} function between two euclidean spaces is a subspace of euclidean space and is also a manifold; and then we showed in \cref{graphs-smooth} that the graph of a \emph{smooth} function is a submanifold of euclidean space. And indeed, the graphs of continuous-but-not-smooth functions provide examples of subspaces that are manifolds but not submanifolds. 

\begin{example}
	Let $f : \R \to \R$ be the function $f(x) = |x|$. Show that the graph of $f$ is not a submanifold of $\R^2$. 
\end{example}

\section{Smooth functions}

We will now define what it means for a function between two manifolds to be smooth. Recall that in \cref{multi} we saw that the case when the codomain is $\R$ was the most important. The same will be true here. You may find that the number of definitions of ``smooth'' gets slightly out of hand; if you do, just remember that, whenever multiple definitions of ``smooth'' make sense, they'll all agree. 

\subsection{Smooth real-valued functions}

\begin{definition}[Smooth real-valued function] \label{smooth-real-valued} \index{smooth}
	Let $X$ be a manifold. A function $f : X \to \R$ is said to be \emph{smooth} if, for any chart $(U, x) \in \mathscr{A}_X$, the function $f \circ x^{-1}$ is smooth.
	\[ \begin{tikzcd} x(U) \ar{r}{x^{-1}} & U \ar{r}{f} & \R \end{tikzcd} \]
\end{definition}

There are two disconcerting things about this definition. First, there is the issue that checking smoothness of a particular function seems next to impossible: the maximal atlas $\mathscr{A}_X$ will have \emph{many, many} charts, some that we may not even know about, and the definition makes it seem like we need to check smoothness on all of them! Thankfully, we don't need to check on all possible charts. As soon as we check smoothness on any atlas, even if it's not maximal, we're done. 

\begin{lemma} \label{smooth-real-valued-local}
	Let $X$ be a manifold and $f : X \to \R$ a function. Suppose that there exists an atlas $\mathscr{A} \subseteq \mathscr{A}_X$ such that $f \circ x^{-1}$ is smooth for any $(U, x) \in \mathscr{A}$. Then $f$ is smooth. 
\end{lemma}

\begin{proof}
	Suppose $(V, y) \in \mathscr{A}_X$. For any $(U, x) \in \mathscr{A}$, we have maps as follows.
	\[ \begin{tikzcd} y(U \cap V) \ar{r}{y^{-1}} \ar{d}[swap]{x \circ y^{-1}} & U \cap V \ar{r}{f} & \R \\ x(U \cap V) \ar[bend right]{ur}[swap]{x^{-1}} \end{tikzcd} \]
	Since $(V, y)$ and $(U, x)$ are both elements of the maximal atlas $\mathscr{A}_X$, we know that the transition function $x \circ y^{-1}$ is smooth. Since $(U, x) \in \mathscr{A}$, we know by assumption that $f \circ x^{-1}$ is smooth. Thus \[ f \circ y^{-1} = (f \circ x^{-1}) \circ (x \circ y^{-1}) \]
	is a smooth function $y(U \cap V) \to \R$. 
	
	Now note that, since $\mathscr{A}$ is an atlas, its charts cover all of $X$, so
	\[ \bigcup_{(U,x) \in \mathscr{A}} y(U \cap V) = y(V). \]
	Since $f \circ y^{-1}$ is smooth when restricted to each the open subset $y(U \cap V)$ for all $(U, x) \in \mathscr{A}$, we conclude that $f \circ y^{-1}$ must be smooth on all of $y(V)$.
\end{proof}

The second disconcerting thing is the following. Suppose $U \subseteq \R^m$ is an open subset and $f : U \to \R$ is a function. On the one hand, we defined what it means for $f$ to be smooth back in \cref{smooth-multi}. On the other hand, we can regard $U$ as an open submanifold of $\R^n$ (cf. \cref{open-submanifold}) and then \cref{smooth-real-valued} gives us another definition of what it means for $f$ to be smooth. A priori, it could be that these definitions conflict. Thankfully, they do not. 

\begin{exercise}
	Suppose $U \subseteq \R^m$ is an open subset and $f : U \to \R$ is a function. Show that $f$ is smooth in the sense of \cref{smooth-multi} if and only if it is smooth in the sense of \cref{smooth-real-valued}. 
	\begin{hint}
		You might find \cref{smooth-real-valued-local} useful. 
	\end{hint}
\end{exercise}

From here on out, we'll use the fact that these two definitions of smoothness coincide without comment. 

\subsubsection*{Combining smooth real-valued functions}

\begin{definition}
	If $X$ is a manifold, we write $\O(X)$ for the set of smooth functions $X \to \R$. 
\end{definition}

Here is the standard set of stability properties we would want. 

\begin{exercise} \label{smooth-real-valued-stable}
	\begin{enumerate}[(a)]
		\item Show that $\O(X)$ is a vector space. 
		\item If $f, g \in \O(X)$, show that $fg \in \O(X)$ also. 
		\item If $f \in \O(X)$ and $f(x) \neq 0$ for all $x$, show that $1/f \in \O(X)$ also. 
		\begin{hint}
			You might decide to do part (d) first, and then use it to prove (c). 
		\end{hint}
		\item Suppose $f \in \O(X)$, $U$ is an open subset of $\R$, $g : U \to \R$ is smooth, and $f(X) \subseteq U$. Show that $g \circ f \in \O(X)$. 
	\end{enumerate}
\end{exercise}

\begin{unimportantremark}
	The first two stability properties in \cref{smooth-real-valued-stable} show that $\O(X)$ is an $\R$-algebra. 
\end{unimportantremark}

\begin{proposition} \label{preponderance}
	Suppose $K$ is a compact subset of $X$ and $U$ is an open neighborhood of $K$. Then there exists a smooth function $f \in \O(X)$ such that $0 \leq f(x) \leq 1$ for all $x \in \R$, $f(x) = 1$ for all $x \in K$, and $f(x) = 0$ for all $x \notin U$. 
\end{proposition}

\begin{proof}
	Suppose $(V, x)$ is a chart containing some $p \in K$. Replacing $x$ with $x - x(p)$, we can assume that $x(p) = 0$. By shrinking $V$, we can assume that $\bar{V} \subseteq U$ (cf. \cref{preponderance-details}). Then $x(V)$ is an open neighborhood of $0 = x(p) \in \R^n$, so there exists $\epsilon > 0$ such that $B(0, \epsilon) \subseteq x(V)$. Let $\psi : \R^n \to \R$ be a bump function supported on $\overline{B}(0, \epsilon)$ (cf. \cref{multivariable-bump-round-support}). Then $\psi \circ x$ is a smooth function on $V$ which is strictly positive on an open neighborhood of $p$ contained entirely inside $V$ (namely, $x^{-1}(B(0,\epsilon))$), and zero elsewhere. Define $f_p : X \to \R$ by
	\[ f_p(x) = \begin{cases} \psi \circ x & \text{if } x \in V \\ 0 &  \text{if } x \notin V. \end{cases} \]
	Then $f_p$ is smooth, and it is strictly positive on an open neighborhood $V_p$ of $p$ contained inside $U$.  
	
	We can construct such a function $f_p$ for every $p \in K$. Since $K$ is compact, there exist finitely many $p_1, \dotsc, p_n \in K$ such that $K \subseteq V_{p_1} \cup \dotsb \cup V_{p_n}$. Thus $f_{p_1} + \dotsb + f_{p_n}$ is strictly positive on $K$, and zero outside $U$. Since $K$ is compact, there exists $\delta > 0$ such that $f_{p_1} + \dotsb + f_{p_n} \geq \delta$ on $K$. Let $g$ be a bridging function such that $g(x) = 1$ for $x \geq \delta$ and $g(x) = 0$ for $x \leq 0$ (cf. \cref{bridge-function}), and then set $f = g \circ (f_{p_1} + \dotsb + f_{p_n})$.
\end{proof}

\begin{exercise} \label{preponderance-details}
	\begin{enumerate}[(a)]
		\item Prove that we can ``shrink $V$ in order to assume that $\bar{V} \subseteq U$.''
		\item Modify the bump function of \cref{multivariable-bump-round-support} so that it has support $\overline{B}(0, \epsilon)$. 
		\item Explain why $\psi \circ x$ and $f_p$ are smooth. 
	\end{enumerate}
\end{exercise}

% TODO: Partitions of unity

\subsection{Smooth functions between manifolds}

Here is a definition that looks a bit terrifying on the face of it. 

\begin{definition}[Smoothness] \label{smooth-between-manifolds} \index{smooth}
	Suppose $X$ and $Y$ are manifolds and $f : X \to Y$ is a function. Then $f$ is \emph{smooth} if, for every chart $(V, y) \in \mathscr{A}_Y$ and every chart $(U, x) \in \mathscr{A}_X$ such that $f(U) \subseteq V$, the function $y \circ f \circ x^{-1} : x(U) \to y(V)$ is smooth. 
	\[ \begin{tikzcd} X \ar{rr}{f} && Y \\ 
	U \ar{rr}{f} \ar[hookrightarrow]{u} \ar{d}[swap]{x} &&  V \ar[hookrightarrow]{u} \ar{d}{y} \\ 
	x(U) \ar{rr}{y \circ f \circ x^{-1}} \ar[hookrightarrow]{d} && y(V) \ar[hookrightarrow]{d} \\ 
	\R^m && \R^n \end{tikzcd} \]
\end{definition}

Once again, we don't need to check all possible charts. The statement is a bit complicated, but it's exactly the statement you might hope for.  

\begin{exercise} \label{smooth-local}
	Suppose $X$ and $Y$ are manifolds and $f : X \to Y$ is a function. Suppose further that there exists an atlas $\mathscr{A} \subseteq \mathscr{A}_Y$ such that, for every $(V, y) \in \mathscr{A}$, there exists an atlas $\mathscr{A}' \subseteq \mathscr{A}_{f^{-1}(V)}$, such that for every $(U, x) \in \mathscr{A}'$, the function $y \circ f \circ x^{-1} : x(U) \to y(V)$ is smooth. Show that $f$ is smooth.
\end{exercise}

We should also note that we haven't redefined the word ``smooth'' in any conflicting ways. More precisely, we have the following two exercises. 

\begin{exercise}
	If $U \subseteq \R^m$ is an open subset and $f : U \to \R^n$ is a function, show that $f$ is smooth in the sense of \cref{smooth-multi} if and only if it is smooth in the sense of \cref{smooth-between-manifolds}.
\end{exercise}

\begin{exercise}
	Show that a function $f : X \to \R$ is smooth in the sense of \cref{smooth-real-valued} if and only if it is smooth in the sense of \cref{smooth-between-manifolds}.
\end{exercise}

From here on out, we'll use the fact that all of these definitions of smoothness coincide without comment. 

\subsubsection*{Combining smooth maps}

It doesn't make sense to add or multiply maps that take values in a manifold. The only thing that makes sense is composition, and indeed, we can compose. 

\begin{exercise} \label{smooth-stable-composite}
	Suppose $X, Y, Z$ are all manifolds and $f : X \to Y$ and $g : Y \to Z$ are smooth. Show that $g \circ f$ is also smooth. 
\end{exercise}  

\subsubsection*{Reduction to smooth real-valued functions}

Finally, we show that the case when the codomain is $\R$ is the most important case. We do this in two steps. 

\begin{proposition} \label{euclidean-valued-smooth-iff-components-smooth}
	Suppose $X$ is a manifold and $f : X \to \R^n$ is a function. Then $f$ is smooth if and only if the component functions $f_i = \pi_i \circ f$ are smooth for all $i = 1, \dotsc, n$. 
\end{proposition}

\begin{proof}
	The single chart $(\R^n, \id)$ is a sub-atlas of the euclidean atlas on $\R^n$, so \cref{smooth-local} says that $f$ is smooth if and only if, for every $(U, x) \in \mathscr{A}_X$, the composite $f \circ x^{-1} : x(U) \to \R^n$ is smooth. This is a function from an open subset of euclidean space into euclidean space, so we know from the definition of smoothness in \cref{multi} that $f \circ x^{-1}$ is smooth if and only if its component functions $\pi_i \circ (f \circ x^{-1})$ are smooth for all $i$. But 
	\[ \pi_i \circ (f \circ x^{-1}) = (\pi_i \circ f) \circ x^{-1} = f_i \circ x^{-1}, \]
	and $f_i \circ x^{-1}$ is smooth for all $(U, x) \in \mathscr{A}_X$ if and only if $f_i$ is smooth, again by \cref{smooth-local}. 
\end{proof}

\begin{proposition} \label{smooth-iff-real-valued-pullbacks-smooth}
	Suppose $f : X \to Y$ is a function between two manifolds. Then the following are equivalent. 
	\begin{enumerate}[(a)]
		\item $f$ is smooth. 
		\item For every open subset $V \subseteq Y$ and every smooth function $g :  V \to \R$, the composite $g \circ f|_{f^{-1}(V)}$ is a smooth function $f^{-1}(V) \to \R$. 
	\end{enumerate}
\end{proposition}

\begin{proof}
	One direction follows immediately from \cref{smooth-stable-composite}. Conversely, suppose $f$ has the property described in (b). Choose a chart $(V, y) \in \mathscr{A}_Y$ of dimension $n$. By assumption, $y_i \circ f|_{f^{-1}(V)}$ is a smooth function $f^{-1}(V) \to \R$ for all $i = 1, \dotsc, n$. By \cref{euclidean-valued-smooth-iff-components-smooth}, this means that $y \circ f|_{f^{-1}(V)}$ is smooth. Since this is true for all charts $(V, y) \in \mathscr{A}_Y$, we conclude that $f$ is smooth.  
\end{proof}

\begin{unimportantremark}
	Suppose $f : X \to Y$ is a smooth function between manifolds. For any $g \in \O(Y)$, we have $g \circ f \in \O(X)$. Thus $g \mapsto g \circ f$ defines a function $f^\sharp : \O(Y) \to \O(X)$, which is actually a homomorphism of $\R$-algebras. 
\end{unimportantremark}

\subsection{Paths}

{\color{blue} Incomplete}

% TODO: Paths

\subsection{Diffeomorphisms}

Diffeomorphisms are how we formalize the idea that two manifolds are ``basically the same.''

\begin{definition}[Diffeomorphism] \index{diffeomorphism}
	Suppose $X$ and $Y$ are manifolds. A function $f : X \to Y$ is a \emph{diffeomorphism} if it is smooth, bijective, and the inverse function $f^{-1} : Y \to X$ is also smooth. If there exists a diffeomorphism $f : X \to Y$, then $X$ and $Y$ are \emph{diffeomorphic}. 
\end{definition}

\begin{exercise}
	Regard $\R$ as a manifold in two different ways: once with the euclidean atlas, and next with the maximal atlas containing $(\R, x \mapsto x^3)$. Show that these two manifolds are diffeomorphic. 
\end{exercise}

\begin{exercise}
	If $X$ is a manifold and $(U, x) \in \mathscr{A}_X$, prove that $x : U \to x(U)$ is a diffeomorphism. 
\end{exercise}

\begin{definition}[Local diffeomorphism] \index{local diffeomorphism|see {diffeomorphism, local diffeomorphism}} \index{diffeomorphism!local diffeomorphism}
	Suppose $X$ and $Y$ are manifolds. A function $f : X \to Y$ is a \emph{local diffeomorphism} at a point $p \in X$ if there exists an open neighborhood $U$ of $p$ such that $f|_U$ is a diffeomorphism $U \to f(U)$. 
\end{definition}

\begin{exercise}
	Consider the function $f : \R \to S^1$ given by $f(t) = (\cos(t), \sin(t))$. Show that $f$ is a local diffeomorphism. 
\end{exercise}

% TODO: Pictures of the universal cover of the circle