\chapter{Preliminaries} \label{prerequisites}

The sections in this chapter cover some ideas relating to point-set topology and linear algebra that will be invoked in the main part of the text. Some sections are merely intended to establish notation; others cover topics that you may not have encountered before. I encourage skipping this chapter, and referring back to it only when you need to. 

You should consult a dedicated linear algebra textbook to review the definitions of vector spaces, subspaces, linear independence, span, dimension, linear maps, matrices, matrix multiplication, determinants, and minors. You should also consult an analysis textbook to review the definitions of metric spaces, equivalence of metrics, open subsets, closed subsets, continuous maps, compactness, and connectedness.

\section{Little-oh notation} \label{little-oh}

Suppose $X$ is a metric space\footnotemark\ and $x_0 \in X$ is a point. 

\footnotetext{Even more generally, $X$ could be a topological space here. See \cref{topological-spaces}.}

\begin{definition}
	A function $g : X \setminus \{x_0\} \to \R$ is \emph{positive} if $g(x) > 0$ for all $x \in X \setminus \{x_0\}$. Similarly, $g$ is \emph{non-negative} if $g(x) \geq 0$ for all $x \in X \setminus \{x_0\}$.
\end{definition}

Suppose $g$ is a positive function. \Cref{little-o-definition} below formulates what it means for a non-negative function $f : X \setminus \{x_0\} \to \R$ to be ``little-oh of $g$ as $x$ tends to $x_0$.'' Intuitively, this condition means that $f$ is \emph{much smaller} than $g$ near $x_0$. 

\begin{definition} \label{little-o-definition}
	A function $f : X \setminus \{x_0\} \to \R$ is \emph{little-oh of $g$ as $x$ tends to $x_0$}, written \[ f(x) = o(g(x)) \text{ as } x \to x_0 \] 
	if, for every $\epsilon > 0$, there exists an open neighborhood $U$ of $x_0$ such that $f(x) \leq \epsilon g(x)$ for all $x \in U \setminus \{x_0\}$. Equivalently, this means that 
	\[ \lim_{x \to x_0} \frac{|f(x)|}{g(x)} = 0. \]
	When $x_0$ can be inferred from context, we write simply $f = o(g)$. 
\end{definition} 

\begin{remark}
	It's worth noting that the use of the symbol ``$=$'' above is mathematically abusive. The left-hand side of the ``$=$'' is a function and the right-hand side is a property of functions; of course, a function cannot literally be equal to a property. Rather, the ``$=$'' is being used here to mean that the function on the left-hand side \emph{has} the property described on the right-hand side. As annoying as it is, this abuse of notation is fairly standard, so it's probably best to just get used to it. 
\end{remark}

\begin{exercise} \label{little-o-equivalent-exercise}
	Prove that the two conditions in \cref{little-o-definition} are in fact equivalent. 
\end{exercise}

\begin{exercise}
	Let $X = \R$ and $x_0 = 0$. For each of the functions $f$ and $g$ described below, sketch graphs of $f$ and $g$ and then determine whether or not $f = o(g)$ as $x \to 0$.
	\begin{enumerate}[(a)]
		\item $f(x) = x$ and $g(x) = x^2$. 
		\item $f(x) = x$ and $g(x) = |x|$. 
		\item $f(x) = x^2$ and $g(x) = |x|$. 
	\end{enumerate}
\end{exercise}

\begin{exercise} \label{little-o-vector-space}
	For a fixed positive function $g : X \setminus \{x_0\} \to \R$, prove that the set of functions $f : X \setminus \{x_0\} \to \R$ which are $o(g)$ as $x \to x_0$ is a vector space under the natural operations. In other words, verify the following three facts. 
	\begin{enumerate}[(1)]
		\item (``Zero is small'') The zero function is $o(g(x))$. 
		\item (``Scalar multiples of small are still small'') If $c \in \R$ is a scalar and $f(x) = o(g(x))$, then 
		\[ cf(x) = o(g(x)). \] 
		\item (``Sum of smalls is still small'') If $f_1(x) = o(g(x))$ and $f_2(x) = o(g(x))$, then \[ (f_1 + f_2)(x) = o(g(x)). \]
	\end{enumerate}
\end{exercise}

\begin{exercise}[``Smaller than small is still small''] \label{less-than-small-is-small}
	Suppose $f_1, f_2 : X \setminus \{x_0\} \to \R$ are functions such that $|f_1(x)| \leq |f_2(x)|$ for all $x$ in a punctured neighborhood of $x_0$, and $f_2(x) = o(g(x))$ as $x \to x_0$. Then $f_1(x) = o(g(x))$ also. 
\end{exercise}

\begin{comment} 	
\begin{exercise} \label{little-o-multiplicative}
Suppose $f_1, f_2, g_1, g_2 : X \setminus \{x_0\} \to \R$ are functions with $g_1, g_2$ positive such that $f_1(x) = o(g_1(x))$ and $f_2(x) = o(g_2(x))$ as $x \to x_0$. Prove that $(f_1 \cdot f_2)(x) = o((g_1 \cdot g_2)(x))$.
\end{exercise} 

\begin{exercise}
Suppose $f : X \to \R$ is continuous at $x_0$ and $f(x_0) = 0$, and that $f = o(g)$ as $x \to  x_0$ for some function $g : X \setminus\{x_0\} \to \R$. Suppose further that $s : \R \to \R$ is a function such that $s = o(|r|)$ as $r \to 0$. Then $s \circ f = o(g)$ as $x \to x_0$. 
\end{exercise}

\begin{proof}
Suppose $\epsilon > 0$. We want to find an open neighborhood $U$ of $x_0$ such that
\[ |s(f(x))| \leq \epsilon|t(g(x))| \]
for $x \in U$. 

Since $s$ is $o(t)$ as $r \to 0$, we know that for any $\epsilon_1 > 0$ there exists $\delta_1 > 0$ such that $|r| < \delta_1$ implies that
\[ \begin{aligned} |s(r)| \leq \epsilon_1|t(r)|. \end{aligned} \] 
Since $f(x_0) = 0$ and $f$ is continuous at $x_0$, there exists a neighborhood $U_1$ of $x_0$ such that
\[ |f(x)| < \delta_1  \]
for all $x \in U_1$. This means that for $x \in U_1$, we have
\[ |s(f(x))| \leq \epsilon_1|t(f(x))|. \]
Since $f = o(g)$ as $x \to x_0$, for any $\epsilon_2 > 0$ there is another neighborhood $U_2$ of $x_0$ such that $|f(x)| \leq \epsilon_2 |g(x)|$. This means that
\[ |s(f(x))| \leq \epsilon_1 \epsilon_2 |g(x)|. \]

there then exists $\delta_2 > 0$ such that $|h| < \delta$ implies that $|df_a(h) + r(h)| < \delta_1$. Then for $|h| < \delta_2$, we have
\[ |s(df_a(h) + r(h)| \leq \epsilon_1 |df_a(h) + r(h)| 
\leq \epsilon_1|f'(a)||h| + \epsilon_1|r(h)| \]
Since $r$ is $o(|h|)$, for any $\epsilon_2 > 0$ there exists $\delta_3 > 0$ such that $|h| < \delta_3$ implies that $|r(h)| \leq \epsilon_2 |h|$. Let $\delta = \min\{\delta_2, \delta_3\}$. Then for $|h| < \delta$, we can continue the above chain of inequalities to obtain the following.  
\[ |s(df_a(h) + r(h)| \leq \epsilon_1|f'(a)||h| + \epsilon_1 \epsilon_2|h| = \epsilon_1(|f'(a)| + \epsilon_2)|h| \]
So we let $\epsilon_2 = 1$ and then let 
\[ \epsilon_1 = \frac{\epsilon}{|f'(a)| + 1}. \]
The $\delta$ obtained by going through the above process then has the desired property.
\end{proof}
\end{comment}

% TODO: exercise about composition? i'm not sure how to formuulate the correct statement though


\section{Product metric} \label{product-metric}

\begin{exercise}
	Suppose $X$ and $Y$ are metric spaces with metrics $d_X$ and $d_Y$, respectively. Define a function $d : (X \times Y) \times (X \times Y) \to \R$ by 
	\[ d((x_1,y_1),(x_2,y_2)) = \max\{ d_X(x_1,x_2), d_Y(y_1,y_2) \}. \]
	Show that $d$ is a metric on $X \times Y$. 
\end{exercise}

\begin{exercise}
	Suppose $X$ and $Y$ are metric spaces with metrics $d_X$ and $d_Y$, respectively. Define a function $d : (X \times Y) \times (X \times Y) \to \R$ by 
	\[ d((x_1,y_1),(x_2,y_2)) = \sqrt{ d_X(x_1,x_2)^2 + d_Y(y_1,y_2)^2 }. \]
	Show that $d$ is a metric on $X \times Y$, and that it is equivalent to the metric from \cref{product-metric}.
\end{exercise}

\section{Norms on vector spaces}

\begin{definition} \label{norm-definition}
	Let $V$ be a vector space. A \emph{norm} $\|-\|$ on $V$ is a function $V \to \R$ satisfying the following two axioms. 
	\begin{enumerate}[(N1)]
		\item $|v| \geq 0$.
		\item $|v| = 0$ if and only if $v = 0$. 
		\item $|\lambda v| = |\lambda| |v|$ for all $\lambda \in \R$ and $v \in V$. 
		\item $|v + w| \leq |v| + |w|$ for all $v, w \in V$.
	\end{enumerate}
\end{definition}

\subsection{Norms induce metrics}

\begin{exercise} \label{norm-induces-metric}
	Suppose $\|-\|$ is a norm on a vector space $V$. Show that the function $d : V \times V \to \R$ given by $d(v, w) = |v-w|$ is a metric. 
\end{exercise}

\begin{exercise} \label{addition-continuous}
	Suppose $|-|$ is a norm on a vector space $V$. Show that the function $V \times V \to V$ given by $(v, w) \mapsto v+w$ is continuous, where $V \times V$ is regarded as a metric space via one of the metrics defined in \cref{product-metric}. 
\end{exercise}

\begin{exercise} \label{scalar-multiplication-continuous}
	Suppose $\|-\|$ is a norm on a vector space $V$. Show that the function $\R \times V \to V$ given by $(\lambda, v) \mapsto \lambda v$ is continuous, where $\R \times V$ is regarded as a metric space via one of the metrics defined in \cref{product-metric}. 
\end{exercise}

\subsection{Equivalence of norms}

\begin{definition} \label{norm-equivalence-definition}
	Two norms $|-|$ and $|-|'$ on a vector space $V$ are \emph{equivalent} if there exist nonzero constants $C_1$ and $C_2$ such that
	\[ C_1 |v| \leq |v|' \leq C_2 |v| \]
	for all $v \in V$. 
\end{definition}

\begin{exercise}
	If $V$ is a vector space, show that equivalence of norms is an equivalence relation on the set of all norms on $V$. 
\end{exercise}

\begin{exercise} \label{equivalent-norm-equivalent-metric}
	Suppose $V$ is a vector space and $|-|$ and $|-|'$ are two equivalent norms on $V$. Show that the metrics $d$ and $d'$ corresponding to $|-|$ and $|-|'$ (cf. \cref{norm-induces-metric}) are equivalent. 
\end{exercise}

\subsection{Norms on finite dimensional vector spaces}

If $V$ is a finite dimensional vector space, we can construct a number of norms on $V$ by choosing a basis. Of these, two are of key importance. 

\begin{example} \label{l2-norm}
	Suppose $V$ is a finite dimensional vector space, and $v_1, \dotsc, v_n$ is a basis for $V$. The \emph{$L^2$-norm} on $V$ with respect to this basis is defined by
	\[ |a_1v_1 + \dotsb + a_n v_n| = \sqrt{a_1^2 + \dotsb + a_n^2}, \]
	where $a_1, \dotsc, a_n \in \R$. 
	Axioms (N1) through (N3) are straightforward to verify. For the triangle inequality (N4), suppose $v = a_1 v_1 + \dotsb + a_n v_n$ and $w = b_1 v_1 + \dotsb +  b_n v_n$. Then
	\[ \begin{aligned} |v + w|^2 &= (a_1 + b_1)^2 + \dotsb + (a_n + b_n)^2 \\
	&= |v|^2 + 2(a_1b_1 + \dotsb + a_n b_n) +  |w|^2 \\
	&\leq |v|^2 + 2|v||w| + |w|^2 \\
	&= (|v| + |w|)^2, \end{aligned} \]
	where we have used the Cauchy-Schwarz inequality\footnotemark\ for the third step. Taking square roots yields (N4). 
\end{example} 

\footnotetext{The \emph{Cauchy-Schwarz inequality} states that
\[ (a_1 b_1  + \dotsb + a_n b_n)^2 \leq (a_1^2 + \dotsb + a_n^2) (b_1^2 + \dotsb + b_n^2) \]
for all $a_1, \dotsc, a_n, b_1, \dotsc, b_n \in \R$. For a proof, see \cite[theorem 6.1]{protter-morrey}, for instance.}

\begin{example} \label{max-norm}
	Suppose $V$ is a finite dimensional vector space, and $v_1, \dotsc, v_n$ is a basis for $V$. The \emph{$L^\infty$-norm}, or the \emph{max norm}, on $V$ with respect to this basis is defined by
	\[ |a_1v_1 + \dotsb + a_n v_n|_\mx = \max \{ |a_1|, \dotsc, |a_n| \}. \]
	Axioms (N1) through (N3) are straightforward to verify. For the triangle inequality (N4), suppose $v = a_1 v_1 + \dotsb + a_n v_n$ and $w = b_1 v_1 + \dotsb +  b_n v_n$. Then
	\[ \begin{aligned} |v +  w|_\mx &= \max\{ |a_1 + b_1| + \dotsb + |a_n + b_n|  \} \\
	&\leq \max\{ |a_1| + |b_1|, \dotsb, |a_n| + |b_n| \} \\
	&\leq \max\{ |a_1|, \dotsc, |a_n|\} + \max\{|b_1|, \dotsc,  |b_n| \} \\
	&= |v|_\mx +  |w|_\mx \end{aligned} \]
	where we used the triangle inequality for real numbers for the second step. 
\end{example}

The following states that these two norms are actually equivalent.  

\begin{exercise} \label{max-l2}
	Suppose $V$ is a finite dimensional vector space and $v_1, \dotsc, v_n$ is a basis for $V$. Let $|-|$ denote the $L^2$-norm from \cref{l2-norm} and $|-|_\mx$ the max norm from \cref{max-norm}. Show that, for any $v \in V$, we have
	\[ |v|_\mx \leq |v| \leq \sqrt{n} |v|_\mx. \]
\end{exercise} 

In fact, here is vast generalization of \cref{max-l2}. 

\begin{theorem} \label{finite-dim-norm}
	If $V$ is a finite dimensional vector space, all norms on $V$ are equivalent.
\end{theorem}

\section{Euclidean space} \label{euclidean}

For any non-negative integer $n$, we write $\R^n$ to denote the set of lists of $n$ real numbers. We will sometimes write its elements as horizontal lists of numbers separated by commas, as in \[ (h_1, \dotsc, h_n), \] and sometimes as vertical column of numbers, as in \[ \begin{bmatrix} h_1 \\ \vdots \\ h_n \end{bmatrix}.  \]

\subsection{Linear structure on \texorpdfstring{$\R^n$}{Rn}}

Given $v_1, v_2 \in \R^n$, we define their sum $v_1 + v_2$ by adding the entries coordinate-wise. Given $\lambda \in \R$ and $v \in \R$, we define $\lambda v$ by multiplying each entry of $v$ by $\lambda$. This endows $\R^n$ with the structure of a vector space. 

For each $i = 1,  \dotsc, n$, we define the \emph{$i$th standard basis vector}, denoted $e_i$, to be the list whose $i$th entry is 1 and all other entries are 0. 
\[ \begin{aligned} e_1 &= (1, 0, 0, \dotsc, 0, 0) \\
e_2 &= (0, 1, 0, \dotsc, 0, 0) \\
&\enspace \vdots \\
e_n &= (0, 0, 0, \dotsc, 0, 1) \end{aligned} \]
Observe that
\[ (h_1, \dotsc, h_n) = h_1 e_1 + h_2 e_2 + \dotsb + h_n e_n.  \]
Thus the list $e_1, \dotsc, e_n$ is a basis for $\R^n$. 

For any $i = 1, \dotsc, n$, we let $\pi_i : \R^n \to \R$ denote the $i$th \emph{projection map}, given by 
\[ \pi_i(h_1, \dotsc, h_n) = h_i. \]
This is a linear map. 

\subsection{Norms on \texorpdfstring{$\R^n$}{Rn}}

The $L^2$-norm (cf. \cref{l2-norm}) with respect to the standard basis $e_1, \dotsc, e_n$ of $\R^n$ is the usual norm we use on $\R^n$. Explicitly, if $v = (h_1, \dotsc, h_n) \in \R^n$, we define
\[ |v| := \sqrt{h_1^2 + \dotsb + h_n^2}. \]
Unless explicitly specified otherwise, we regard $\R^n$ as a metric space with the metric induced by this norm (cf. \cref{norm-induces-metric}). 

On the other hand, we could also endow $\R^n$ with the max norm (cf. \cref{max-norm}) with respect to the standard basis. We denote this norm by $|-|_\mx$. Explicitly, if $v = (h_1, \dotsc, h_n) \in \R^n$, we have
\[ |v|_\mx = \max\{ |h_1|, \dotsc, |h_n| \}. \]

\section{Matrices}

The following is a whirlwind review of some facts about matrices, mostly intended to establish notation. The reader is expected to remember definitions of matrix multiplication, determinants, and minors. Few proofs are included in this section; for more details, you are encouraged to reference a dedicated linear algebra text.

\begin{definition}
	We write $\mathscr{M}_{n,m}$ for the set of all $n \times m$ matrices (ie, the matrices with $n$ rows and $m$ columns). This is naturally a vector space, where addition and scalar multiplication are defined entrywise.  
\end{definition}

\begin{definition}
	Let $\GL_n$ denote the subset of $\mathscr{M}_{n \times n}$ consisting of invertible $n \times n$ matrices.
\end{definition}

\begin{lemma}
	Let $A$ be a matrix. Then all of the following numbers are equal. 
	\begin{enumerate}[(1)]
		\item The dimension of the span of the columns of $A$. \item The dimension of the span of the rows of $A$. 
		\item The size of largest nonzero minor of $A$. 
	\end{enumerate}
	This integer is called the \emph{rank} of $A$, and is denoted $\rank(A)$. 
\end{lemma}

\subsection{Matrix representations of linear maps} \label{matrix-representations}

\begin{definition}
	If $V$ and $W$ are both vector spaces, we write $\mathscr{L}(V, W)$ for the set of all linear maps $V \to W$. This is naturally a vector space. 
\end{definition}

\begin{definition}
	If $V$ is a vector space, let $\GL(V)$ denote the set of invertible linear maps $V \to V$, regarded as a subset of $\mathscr{L}(V,V)$. 
\end{definition}

Throughout the rest of this section, we assume that $U$, $V$, and $W$ are finite dimensional vector spaces. 

\begin{definition}
	Suppose $B$ denotes a basis $v_1, \dotsc, v_n$ for $V$. Then any $v \in V$ can be written uniquely as $a_1v_1 + \dotsb + a_n v_n$ for some scalars $a_1, \dotsc, a_n \in \R$, and we define the \emph{representation of $v$ with respect to $B$}, denoted $[v]_B$, to be the column vector with $n$ entries that records the scalars $a_1, \dotsc, a_n$. In other words,
	\[ [v]_B = \begin{bmatrix} a_1 \\ \vdots \\ a_n \end{bmatrix}. \]
\end{definition}

\begin{lemma}
	If $B$ denotes a basis $v_1, \dotsc, v_n$ is a basis for $V$, then the function $V \to \R^n$ given by $v \mapsto [v]_B$ is an isomorphism, with inverse given by
	\[ \begin{bmatrix} a_1 \\ \vdots \\ a_n \end{bmatrix} \mapsto a_1 v_1 + \dotsb + a_n v_n. \]
\end{lemma}

\begin{definition} \label{matrix-representation-linear}
	Suppose $\ell : V \to W$ is a linear map. Let $B$ and $C$ denote bases $v_1, \dotsc, v_m$ and $w_1, \dotsc, w_n$ for $V$ and $W$, respectively. The \emph{matrix representation of $\ell$ with respect to $B$ and $C$}, denoted $[\ell]_{B,C}$, is the $n \times m$ matrix whose $i$th column of $[\ell]_{B,C}$ is $[\ell(v_i)]_C$. In other words,
	\[ [\ell]_{B,C} = \begin{bmatrix} [\ell(v_1)]_C & \dotsb & [\ell(v_m)]_C \end{bmatrix}. \]
\end{definition}

\begin{lemma} \label{matrix-rep-isomorphism}
	Suppose that $V$ and $W$ are $m$- and $n$-dimensional vector spaces, respectively, and that and $B$ and $C$ are bases for $V$ and $W$, respectively. Then the function $\ell \mapsto [\ell]_{B,C}$ is an isomorphism \[ \begin{tikzcd} \mathscr{L}(V, W) \ar{r} & \mathscr{M}_{n,m}. \end{tikzcd} \] 
\end{lemma}

\begin{lemma}
	If $\ell : V \to W$ is a linear map between finite dimensional vector spaces and $B$ and $C$ are bases for $V$ and $W$, respectively, then \[ [\ell]_{B,C}[v]_B = [\ell(v)]_C \]
	for any $v  \in V$. 
\end{lemma}

\begin{lemma}
	If $\ell : U \to V$ and $\ell' :  V \to W$ are linear maps, and $A, B$ and $C$ are bases for $U, V$, and $W$, respectively, then 
	\[ [\ell' \circ \ell]_{A,C} = [\ell']_{B,C}[\ell]_{A,B}. \]
\end{lemma}

\begin{definition}
	If $\ell : V \to W$ is a linear map, then the \emph{rank} of $\ell$, denoted $\rank(\ell)$ is defined to be the dimension of the range of $\ell$. 
\end{definition}

\begin{lemma}
	If $\ell : V \to W$ is a linear map and $B$ and $C$ are bases for $V$ and $W$, respectively, then \[ \rank (\ell) = \rank [\ell]_{B,C}. \]
\end{lemma}

\begin{definition} \label{matrix-representation-standard}
	If $\ell : \R^m \to \R^n$ is a linear map, the \emph{standard matrix representation} of $\ell$, denoted $[\ell]$, is the matrix of $\ell$ with respect to the standard bases on $\R^m$ and $\R^n$.
\end{definition} 

\begin{lemma}
	If $A$ is a $n \times m$ matrix and $\ell_A : \R^m \to \R^n$ is the linear map $\ell_A(v) = Av$, then \[ [\ell_A] = A. \]
\end{lemma}

\subsection{Matrix norm} \label{matrix-norm}

Observe that $\mathscr{M}_{n \times m}$ is a finite dimensional vector space. If $E_{j,i}$ denotes the matrix whose $(j,i)$-entry is 1 and all other entries are 0 for $i = 1, \dotsc, m$ and $j = 1, \dotsc, n$, then the list of all of these $mn$ matrices forms a basis for $\mathscr{M}_{n \times m}$. We can use this basis to define a $L^2$ and max norm on $\mathscr{M}_{n \times m}$, as in \cref{l2-norm,max-norm}. Explicitly, if $A$ is a $n \times m$ matrix whose $(j,i)$-entry is $a_{j,i}$, we have the following. 
\[ \begin{aligned} \|A\| &= \sqrt{ \sum_{j,i} a_{j,i}^2 } \\
\|A\|_\mx &= \max_{j,i} |a_{j,i}| \end{aligned}  \]
We know from \cref{max-l2} that these two norms are equivalent. All of the following statements are independent of which norm on $\mathscr{M}_{n \times m}$ we have in mind. 

\begin{lemma}
	The function $\det : \mathscr{M}_{n \times n} \to \R$ is continuous.
\end{lemma}

\begin{proof}
	Determinants can be computed using cofactor expansions, so $\det A$ is a polynomial in the entries of $A$.
\end{proof}

\begin{corollary} \label{gln-open}
	$\GL_n$ is an open subset of $\mathscr{M}_{n \times n}$.
\end{corollary}

\begin{proof}
	Since $\det : \mathscr{M}_{n \times n} \to \R$ is continuous, and $\R \setminus \{0\}$ is an open subset of the codomain, we see that $\GL_n = \text{det}^{-1}(\R \setminus \{0\})$ must be open in $\mathscr{M}_{n \times n}$. 
\end{proof}

\begin{lemma} \label{inversion-homeomorphism}
	The function $\GL_n \to \GL_n$ given by $A \mapsto A^{-1}$ is a homeomorphism. 
\end{lemma}

\begin{proof}
	This function is its own inverse, so it is sufficient to show that $A \mapsto A^{-1}$ is continuous. But \[ A^{-1} = \det(A)^{-1} \adj(A), \]
	where $\adj(A)$ denotes the adjugate matrix of $A$. 
	We know that $\det(A)$ is a polynomial in the entries of $A$. Each entry of $\adj(A)$ is a minor of $A$ and is therefore also a polynomial in the entries of $A$. The lemma follows. 
\end{proof}

\section{Operator norm} \label{operator-norm}

\subsection{Basics} \label{operator-norm-basic}

\begin{definition} \label{operator-norm-definition}
	Suppose $\ell : \R^m \to \R^n$ is a linear map. The \emph{operator norm} of $\ell$, denoted $\|\ell\|$, is defined to be
	\[ \|\ell\| = \sup\, \{ |\ell(v)| : |v| \leq 1 \}. \]
\end{definition} 

The operator norm $\|\ell\|$ is \emph{not} the same as the norm of the standard matrix representation $\|[\ell]\|$ as defined in \cref{matrix-norm}. The following gives an important reformulation of \cref{operator-norm-definition}.  

\begin{exercise} \label{operator-norm-reformulation}
	Suppose $\ell : \R^m \to \R^n$ is a linear map. Show that 
	\[ \|\ell\| = \inf \, \{ \lambda \in \R : |\ell(v)| \leq \lambda |v| \text{ for all } v \in \R^m \}. \]
	In particular, this means that $|\ell(v)| \leq \|\ell\| \cdot |v|$ for all $v \in \R^m$. 
\end{exercise}

Here are some fundamental properties of the operator norm. 

\begin{exercise} \label{operator-norm-is-norm}
	Show that the operator norm is a norm on $\mathscr{L}(\R^m, \R^n)$ in the sense of \cref{norm-definition}. 
\end{exercise}

\begin{lemma} \label{norm-is-realized}
	Suppose $\ell : \R^m \to \R^n$ is a nonzero linear map. Then there exists $v_0 \in \R^m$ such that $|v_0| = 1$ and $|\ell(v_0)| = \|\ell\|$. 
\end{lemma}

\begin{proof}
	The ``unit disk'' \[ D = \{ x \in \R^m : |x| \leq 1 \} \]
	is a nonempty, connected, and compact subset of $\R^m$. The function $v \mapsto |\ell(v)|$ is also a continuous function $D \to \R$, so its image $|\ell(D)|$ must also be a nonempty, connected, and compact subset of $\R$. In other words, it must be a closed interval in $\R$. This means that there exists $v_0 \in D$ such that 
	\[ |\ell(v_0)| = \sup |\ell(D)| = \|\ell\|. \]
	Since $\ell \neq 0$, we know $\|\ell\| \neq 0$ from \cref{operator-norm-is-norm}, which in turn means that $v_0 \neq 0$. To prove that $|v_0| = 1$, assume for a contradiction that $|v_0| < 1$, and let $c = 1/|v_0| > 1$. Then $|cv_0| = |c||v_0| = 1$, so $cv_0 \in D$. But then
	\[ |\ell(cv_0)| = |c\ell(v_0)| = c|\ell(v_0)| > |\ell(v_0)| = \sup |\ell(D)|, \]
	which is absurd. 
\end{proof}

\begin{lemma} \label{linear-and-small}
	Suppose $\ell : \R^m \to \R^n$ is a linear map and $|\ell(v)| = o(|v|)$ as $v \to 0$. Then $\ell = 0$. 
\end{lemma}

\begin{proof}
The fact that $|\ell(v)| = o(|v|)$ tells us that
\begin{equation} \label{big-sequence} \lim_{v \to 0} \frac{|\ell(v)|}{|v|} = 0. \end{equation}
Suppose for a contradiction that $\ell \neq 0$. Using \cref{norm-is-realized}, choose $v_0 \in \R^m$ such that $|v_0| = 1$ and $|\ell(v_0)| = \|\ell\|$. For scalars $c$, observe that $cv_0 \to 0$ as $c \to 0$, which means that \cref{big-sequence} implies that
\[ \lim_{c \to 0} \frac{|\ell(cv_0)|}{|cv_0|} = 0.  \]
But observe that, for nonzero $c$, we have
\[ \frac{|\ell(cv_0)|}{|cv_0|} = \frac{|c\ell(v_0)|}{|cv_0|} = \|\ell\| \]
so
\[ 0 = \lim_{c \to 0} \frac{|\ell(cv_0)|}{|cv_0|} = \lim_{c \to 0} \|\ell\| = \|\ell\|. \]
This contradicts \cref{operator-norm-is-norm}. 
\end{proof}

\subsection{Operator norm as a metric \texorpdfstring{$\star$}{*}}

Since the operator norm is a norm by \cref{operator-norm-is-norm}, we can regard $\mathscr{L}(\R^m, \R^n)$ as a metric space by \cref{norm-induces-metric}. It turns out that many natural maps are continuous when we do this. For example, we have already seen that addition and scalar multiplication are continuous in \cref{addition-continuous,scalar-multiplication-continuous}. 

\begin{exercise}[``Evaluation is continuous''] \label{evaluation-continuous}
	Suppose $v_0 \in \R^m$ is a fixed vector. Then the ``evaluate at $v_0$'' function $\mathscr{L}(\R^m, \R^n) \to \R^n$ given by $\ell \mapsto \ell(v_0)$ is continuous. \begin{hint} Since the ``evaluate at $v_0$'' function is linear, it is sufficient to check continuity at 0. \end{hint}
\end{exercise}

\begin{exercise}[Submultiplicativity] \label{operator-norm-submultiplicative}
	Suppose $\ell : \R^m \to \R^n$ and $\ell' : \R^n \to \R^p$ are both linear maps. Show that \[ \|\ell' \circ \ell\| \leq \|\ell'\| \cdot \|\ell\|. \]
	Give an example to show that this inequality can be strict. 
\end{exercise}

\begin{exercise}[``Composition is continuous''] \label{composition-continuous}
	Show that the map
	\[ \begin{tikzcd} \mathscr{L}(\R^n, \R^p) \times \mathscr{L}(\R^m, \R^n) \ar{r} & \mathscr{L}(\R^m, \R^p) \end{tikzcd} \]
	given by $(\ell', \ell) \mapsto \ell' \circ \ell$ is continuous.
\end{exercise}

\begin{exercise}
	Suppose $\ell : \R^m \to \R^n$ is linear and $A = [\ell]$ is its standard matrix representation. Prove the following. 
	\begin{enumerate}[(a)]
		\item $\|\ell\| \leq \|A\| \leq \sqrt{n} \|\ell\|$. 
		\item $\|A \|_\mx \leq \|\ell\| \leq \sqrt{mn} \|A\|_\mx$. 
	\end{enumerate}
\end{exercise}

% \cite[section 2.3.2]{golub}

\begin{exercise} \label{matrix-to-maps-homeomorphism}
	Show that the isomorphism $\mathscr{L}(\R^m, \R^n) \to \mathscr{M}_{n \times m}$ of \cref{matrix-rep-isomorphism}, given by $\ell \mapsto [\ell]$, is a homeomorphism. 
\end{exercise}

Under the homeomorphism $\mathscr{L}(\R^n, \R^n) \to \mathscr{M}_{n \times n}$ of \cref{matrix-to-maps-homeomorphism}, the set $\GL(\R^n)$ of invertible linear maps $\R^n \to \R^n$ corresponds to the set $\GL_n$ of invertible $n \times n$ matrices; since $\GL_n$ is an open subset of $\mathscr{M}_{n \times n}$ by  \cref{gln-open}, we conclude that $\GL(\R^n)$ is an open subset of $\mathscr{L}(\R^n, \R^n)$. Similarly, the following is a consequence of combining \cref{inversion-homeomorphism,matrix-to-maps-homeomorphism}. 

\begin{exercise} \label{inversion-homeomorphism-maps}
	Show that the map $\GL(\R^n) \to \GL(\R^n)$ given by $\ell \mapsto  \ell^{-1}$ is a homeomorphism. 
\end{exercise}

\section{Topological spaces} \label{topological-spaces}

This is a bare-bones introduction to topological spaces, intended for readers who have seen metric spaces but not topological spaces before. 

\subsection{Basics} \label{topological-spaces-basics}

\begin{definition}
	Let $X$ be a set. A \emph{topology} on $X$ is a set $\tau_X$ of subsets of $X$, called \emph{open subsets}, which satisfy the following axioms.
	\begin{enumerate}[(T1)]
		\item $\emptyset$ and $X$ are both open subsets. 
		\item The union of an arbitrary collection of open subsets is open. 
		\item The intersection of a finite collection of open subsets is open. 
	\end{enumerate}
	A pair $(X, \tau_X)$ consisting of a set $X$ and a topology $\tau$ on $X$ is called a \emph{topological space}. Often, we write simply ``$X$'' in place of the pair $(X, \tau_X)$. 
\end{definition}

Topological spaces come up throughout mathematics, and in general can be very very bizarre. We will not delve into the general theory of topological spaces, and won't see any of these bizarre examples. Instead, you are encouraged to content yourself with the following examples. 

\begin{example}[Metric spaces] \label{metric-to-topological}
	If $X$ is a metric space, the collection of subsets of $X$ that are open with respect to the metric defines a topology on $X$. Thus, every metric space can be regarded as a topological space in a natural way. It's worth noticing that this process of regarding a metric space as a topological space ``forgets information,'' since equivalent metrics on a set will define the same topology. 
\end{example}

\begin{example}[Quotient spaces] \label{quotient-space}
	Suppose $Y$ is any topological space (for instance, one that comes from a metric space), and let $\sim$ be an equivalence relation on $Y$. For an element $y \in Y$, let $[y]$ denote its equivalence class, and let $X = Y/\sim$ denote the set of all equivalence classes. There is a natural function $\pi : Y \to X$ given by sending a point $y \in Y$ to its equivalence class $[y]$. We define a topology on $X$, called the \emph{quotient topology}, by declaring a subset $U \subseteq X$ to be open if and only if its preimage \[ \pi^{-1}(U) = \{ y \in Y : [y] \in U \} \] is open in $Y$. 
\end{example}

\begin{example}[Product spaces] \label{product-space}
	Suppose $X$ and $Y$ are topological spaces. We define a topology $\tau$ on the cartesian product $X \times Y$ by declaring a subset $U$ to be open if it is a union of sets of the form $V \times W$ where $U$ and $V$ are open subsets of $X$ and $Y$, respectively. This is called the \emph{product topology} on $X \times Y$. 
\end{example}

\begin{example} \label{finite-dim-canonical-topology}
	Suppose $V$ is a finite dimensional vector space. Then all norms on $V$ are equivalent by \cref{finite-dim-norm}, so they all define the same topology on $V$ by \cref{metric-to-topological,equivalent-norm-equivalent-metric}. This topology is called the \emph{canonical topology} on $V$. Unless explicitly specified otherwise, we will always regard finite dimensional vector spaces as topological spaces using the canonical topology. 
\end{example}

We now make a few further definitions that readers are invited to compare against the definitions they are familiar with from their experience with metric spaces. 

\begin{definition}[Closed subsets]
	If $X$ is a topological space, a subset $F$ is \emph{closed} if its complement $U$ is \emph{open}. 
\end{definition}

\begin{definition}[Continuous functions]
	If $X$ and $Y$ are topological spaces, a function $f : X \to Y$ is \emph{continuous} if $f^{-1}(U)$ is an open subset of $X$ whenever $U$ is an open subset of $Y$. 
\end{definition}

\begin{example}
	If $Y$ is a topological space, $\sim$ an equivalence relation on $Y$, and $X = Y/\sim$ the quotient space as in \cref{quotient-space} above, then the map $\pi : Y \to X$ which carries a point $y \in Y$ to its equivalence class is continuous. This follows immediately from the definition of the quotient topology.
\end{example}

\begin{exercise}
	If $f : X \to Y$ and $g : Y \to Z$ are continuous functions between topological spaces, then $g \circ f$ is also continuous. 
\end{exercise}

\begin{example}
	If $V$ is a finite dimensional vector space, then the addition map $V \times V \to V$ and the scalar multiplication map $\R \times V \to V$ are continuous functions (using the product topology of \cref{product-space} on the domains). This statement is precisely the same as \cref{addition-continuous,scalar-multiplication-continuous}. 
\end{example}

\begin{example}
	Suppose $\ell : V \to W$ is a linear map between two finite dimensional vector spaces. Then $\ell$ is automatically continuous. 
\end{example}

\begin{definition}[Homeomorphisms]
	Suppose $X$ and $Y$ are topological spaces and $f : X \to Y$ is a function. Then $f$ is a \emph{homeomorphism} if it is bijective, continuous, and $f^{-1} : Y \to X$ is continuous. More generally, $f$ is a \emph{homeomorphism onto its image} if is injective, continuous, and $f^{-1} : f(X) \to X$ is also continuous. 
\end{definition}

\begin{definition}[Subspace] \label{subspace-topology}
	If $X$ is a topological space and $S$ is a subset, then the \emph{subspace topology} on $S$ is
	\[ \tau_S := \{ S \cap U : U \in \tau_X \}. \]
	Further, $S$ equipped with the subspace topology is called a \emph{subspace} of $X$. 
\end{definition}

\begin{exercise}
	If $X$ is a topological space and $S$ is a subspace, prove that the inclusion map $i : S \to X$ is continuous. 
\end{exercise}

\subsection{Hausdorff spaces \texorpdfstring{$\star$}{*}}

\begin{definition}[Hausdorff]
	A topological space $X$ is \emph{Hausdorff} if, for every pair of distinct points $x, y \in X$, there exist disjoint open subsets $U$ and $V$ containing $x$ and $y$, respectively. 
\end{definition}

\begin{example}
	If $X$ is a metric space, the corresponding topological space as in \cref{metric-to-topological} is automatically Hausdorff. 
\end{example}

However, quotient spaces need not be Hausdorff. 

\begin{exercise}
	Let $Y = \R \times \{\pm 1 \}$ inside $\R^2$, and define an equivalence relation $\sim$ by declaring that $(a, 1) \sim (a,-1)$ for all $a \neq 0$. Show that $X = Y/\sim$ is not Hausdorff. 
\end{exercise}

\subsection{Compact and \texorpdfstring{$\sigma$}{sigma}-compact spaces \texorpdfstring{$\star$}{*}}

\begin{definition}[Open covers]
	An \emph{open cover} of a topological space $X$ is a collection $\mathscr{U}$ of open subsets of $X$ which cover $X$, in the sense that \[ \bigcup_{U \in \mathscr{U}} U = X. \]
	A \emph{subcover} of $\mathscr{U}$ is a subset $\mathscr{U}' \subseteq  \mathscr{U}$ which is itself an open cover. 
\end{definition}

\begin{definition}[Compact]
	A topological space $X$ is \emph{compact} if every open subcover has a finite subcover. 
\end{definition}

\begin{definition}[$\sigma$-compact]
	A topological space $X$ is \emph{$\sigma$-compact} if it is a countable union of compact subspaces. 
\end{definition}

\begin{example}
	$\R^n$ is $\sigma$-compact, since
	\[ \R^n = \bigcup_{k = 1}^\infty [-k,k]^n \]
	and each hypercube $[-k,k]^n$ is compact. 
\end{example}

