\chapter{Single variable derivatives} \label{single-var}

The goal in this chapter is to formalize derivatives of functions whose input is a single real number and whose output is also a single real number. 

\section{Two definitions of the derivative}

\subsection{Slope of the tangent line}

Let $S$ be a subset of $\R$ and $a \in S$ an interior point. We would like to define the tangent line to the graph of a function $f : S \to \R$ at $a$. We know intuitively what ``tangent line'' should mean: a line that ``just barely touches'' the graph of $f$ at the point $(a, f(a))$. But it's not immediately clear how to formalize this intuition.  

One approach to formalization is to notice that our intuitive idea of ``tangent line'' can be approximated by secant lines, and secant lines can be formalized without ambiguity. More precisely, for small values of $h$, the point $a+h$ is close to $a$ and still in $S$ since $a$ is an interior point of $S$. The slope of the secant line passing through the two points $(a, f(a))$ and $(a+h, f(a+h))$ is computed by the quantity
\[ \frac{f(a+h)-f(a)}{h}. \]
This quantity is often called a \emph{difference quotient}. See \cref{figure-difference-quotient} for a geometric explanation of this difference quotient. By letting $h \to 0$, we get closer and closer to what we intuitively understand to be the slope of the tangent line. 

\begin{figure}
	\begin{center}
	\begin{tikzpicture}
	\begin{axis}[
	xmin=-0.5,xmax=3,
	ymin=-0.5,ymax=2.25,
	ytick={0.7, 1.5},
	xtick={1,2},
	xticklabels={$a$, $a+h$},
	yticklabels={$f(a)$,$f(a+h)$},
	axis lines=middle,
	axis line style=lightgray,
	width=0.75\textwidth
	]
	\addplot[black,style=thick,<->] expression[domain=-0.5:2.45,samples=50]{(x*(x^2+1)+5)/10}; 
	
	
	\addplot[red,<->] expression[domain=0.5:2.4,samples=2]{0.8*(x-1)+0.7}; 
	
	
	\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{0.7};
	\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{1.5}; 
	\addplot +[mark=none,lightgray,dotted] coordinates {(1, 0) (1, 2)};
	\addplot +[mark=none,lightgray,dotted] coordinates {(2, 0) (2, 2)};
	
	\addplot[gray,dashed] expression[domain=1:2,samples=2]{0.7}; 
	\addplot +[mark=none,gray,dashed] coordinates {(2, 0.7) (2, 1.5)};
	
	\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:1,0.7) {}; 
	
	\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:2,1.5) {}; 
	\end{axis}
	\end{tikzpicture}
\end{center}
\caption{The graph of a function $f$ is depicted in black. The secant line passing through the two points $(a, f(a))$ and $(a+h, f(a+h))$ is depicted in red. The ``run'' between these points, is $h$. The ``rise'' between these points is $f(a+h)-f(a)$. Thus the slope of the secant line, computed as ``rise over run,'' is precisely the difference quotient $(f(a+h)-f(a))/h$.}  \label{figure-difference-quotient}
\end{figure}

\begin{definition} \label{single-var-derivative-definition}
	A function $f : S \to \R$ is \emph{differentiable} at an interior point $a \in S$ if 
	\[ \lim_{h \to 0} \frac{f(a+h) - f(a)}{h} \]
	exists. If this limit does exist, its value is called the \emph{derivative of $f$ at $a$}. We will usually denote this derivative as $f'(a)$, but sometimes also as one of the following, where $x$ is the variable that is being used to denote the argument to $f$.  
	\[ \left.\frac{df}{dx}\right|_{x = a}, \quad \frac{df}{dx} (a), \dotsc \] 
\end{definition}

\begin{exercise}
	Use the definition of the derivative to determine whether or not each of the following functions $f$ is differentiable at the given point $a$. If it is differentiable, calculate $f'(a)$. Check your answers by making sure they agree with geometric intuition and/or rules you learned during your introductory calculus course. 
	\begin{enumerate}[(a)]
		\item $f(x) = |x|$ at $a = 0$.
		\item $f(x) = x$ at $a = 0$. 
		\item $f(x) = x^2$ at $a = 0$. 
		\item $f(x) = x^2$ at $a = 2$. 
		\item $f(x) = \begin{cases} 1 & x \geq 0 \\ 0 & x < 0 \end{cases}$ at $a = 0$. 
	\end{enumerate}
\end{exercise}

\begin{exercise}[Power rule for positive integer exponents] \label{power-rule-positive}
	Prove that, if $f(x) = x^n$ for a positive integer $n$, then $f'(a) = na^{n-1}$. \textit{Hint.} Recall the binomial theorem, which says that
	\[ (a+h)^n = \sum_{k = 0}^n {n \choose k} a^k h^{n-k}. \]
\end{exercise}

\begin{exercise}[Differentiability implies continuity] \label{differentiable-implies-continuous-single}
	Prove that if $f : S \to \R$ is differentiable at an interior point $a \in U$, then $f$ is also continuous at $a$. 
\end{exercise}

We now make the following definition. 

\begin{definition}
	We say that a function $f : S \to \R$ is \emph{differentiable} if $f$ is continuous on $S$ and differentiable at every interior point $a \in S$. Letting $S^\circ$ denote the interior of $S$, the function $S^\circ \to \R$ defined by $a \mapsto f'(a)$ is called the \emph{derivative} of $f$. We will usually denote this function by $f'$, but occasionally we will also use the notation $df/dx$, where again $x$ is the variable that is being used to denote the argument to $f$. 
\end{definition}

Note that \cref{differentiable-implies-continuous-single} makes the continuity on $S^\circ$ automatic. 

\subsection{Best linear approximation}

Before proceeding further, we reinterpret the definition of the derivative in a way that will be better suited to generalization to the multivariable setting. Suppose a function $f : S \to \R$ is differentiable at an interior point $a \in S$. The tangent line to the graph at the point $(a, f(a))$ is the graph of the function
\[ t(x) = f'(a)(x-a) + f(a). \]
Since this function $t$ is ``close'' to $f$ near $a$, we should have that
\[ f(a+h)-f(a) \approx t(a+h) - t(a) = f'(a)h \]
for small values of $h$. See \cref{figure-linear-approximation}. 

\begin{figure}
	\begin{center}
		\begin{tikzpicture}
		\begin{axis}[
		xmin=-0.5,xmax=3,
		ymin=-0.5,ymax=2.25,
		ytick={0.7, 1.1, 1.5},
		xtick={1,2},
		xticklabels={$a$, $a+h$},
		yticklabels={$f(a)$, $t(a+h)$, $f(a+h)$},
		axis lines=middle,
		axis line style=lightgray,
		width=0.75\textwidth
		]
		\addplot[black,style=thick,<->] expression[domain=-0.5:2.45,samples=50]{(x*(x^2+1)+5)/10}; 
		
		
		\addplot[red,<->] expression[domain=0.5:2.4,samples=2]{0.4*x+0.3}; 
		
		
		\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{0.7};
		\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{1.1};
		\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{1.5}; 
		\addplot +[mark=none,lightgray,dotted] coordinates {(1, 0) (1, 2)};
		\addplot +[mark=none,lightgray,dotted] coordinates {(2, 0) (2, 2)};
		
		\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:1,0.7) {}; 
		\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:2,1.5) {}; 
		\node[color=red,draw=red,circle,fill,inner sep=2pt] at (axis cs:2,1.1) {}; 
		\end{axis}
		\end{tikzpicture}
	\end{center}
	\caption{The graph of a function $f$ is depicted in black, and its tangent line $t$ at $a$ is depicted in red. Then the vertical distance $f(a+h)-f(a)$ is approximated by the vertical distance $t(a+h)-t(a)$. Since the tangent line has slope $f'(a)$, this ``rise'' can be computed as $f'(a)$ times the ``run'' (in other words, $f'(a)h$). As $h$ gets smaller, this approximation gets better.}  \label{figure-linear-approximation}
\end{figure}

Notice that the function $h \mapsto f'(a)h$ is a linear map $\R \to \R$ in the sense of linear algebra. One way to interpret the following result is that the function $h \mapsto f'(a)h$ is the \emph{only} linear map that is a good approximation to $h \mapsto f(a+h)-f(a)$. The precise statement of the result uses little-oh notation; if you have not seen this notation before, see \cref{little-oh}.

\begin{proposition} \label{unique-linear-single-var}
	A function $f : S \to \R$ is differentiable at an interior point $a \in S$ if and only if there exists a linear map $\ell : \R \to \R$ such that
	\[ |f(a+h)-f(a) - \ell(h)| = o(|h|) \text{ as } h \to 0. \]
	Moreover, if there exists such a linear map $\ell$, then $\ell$ is uniquely determined: it must be given by \[ \ell(h) = f'(a)h \]
	for all $h \in \R$. 
\end{proposition}

\begin{proof}
	Let $\ell$ be a linear map $\ell(h) = ch$, where $c$ is some scalar. Observe that
	\[ \begin{aligned} \frac{|f(a+h)-f(a)-\ell(h)|}{|h|} &= \left| \frac{f(a+h)-f(a) - ch}{h} \right| \\
	&= \left| \frac{f(a+h)-f(a)}{h} - c \right|. \end{aligned} \]
	This leads to the following sequence of ``if and only if'' statements. 
	\[ \begin{aligned} |f(a+h)-f(a)-\ell(h)| = o(|h|) &\iff \lim_{h \to 0} \frac{|f(a+h)-f(a)-\ell(h)|}{|h|} = 0 \\
	&\iff \lim_{h \to 0} \left| \frac{f(a+h)-f(a)}{h} - c \right| = 0 \\
	&\iff \lim_{h \to 0} \left( \frac{f(a+h)-f(a)}{h} - c \right) = 0 \\
	&\iff \lim_{h \to 0} \frac{f(a+h)-f(a)}{h} = c \end{aligned} \]
	The statement of the proposition follows. 
\end{proof}

\begin{definition}
	If $f : S \to \R$ is differentiable at an interior point $a \in U$, we define $df_a : \R \to \R$ to be the linear map $df_a(h) = f'(a)h$. 
\end{definition}

In other words, $df_a$ is the only good linear approximation to the function \[ h \mapsto f(a+h)-f(a). \] Being the \emph{only good} linear approximation, it is also the \emph{best} linear approximation, whence the name of this section.

The following is an important idea that we will come up frequently in proofs. 

\begin{remark}[``Error in approximation'' function] \label{error-in-approximation}
	Suppose $f : S \to \R$ is differentiable at an interior point $a \in S$. For sufficiently small values of $h$, we can define the ``error in approximation'' function
	\[ r(h) = f(a+h) - f(a) - df_a(h). \]
	See \cref{figure-error-in-approximation}. 
	\begin{figure}
		\begin{center}
			\begin{tikzpicture}
			\begin{axis}[
			xmin=-0.5,xmax=3,
			ymin=-0.5,ymax=2.25,
			ytick={0.7, 1.1, 1.5},
			xtick={1,2},
			xticklabels={$a$, $a+h$},
			yticklabels={$f(a)$, $f(a)+df_a(h)$, $f(a+h)$},
			axis lines=middle,
			axis line style=lightgray,
			width=0.75\textwidth
			]
			\addplot[black,thick,<->] expression[domain=-0.5:2.45,samples=50]{(x*(x^2+1)+5)/10}; 
			
			
			\addplot[black,<->] expression[domain=0.1:2.4,samples=2]{0.4*x+0.3}; 
			
			
			\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{0.7};
			\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{1.1};
			\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{1.5}; 
			\addplot +[mark=none,lightgray,dotted] coordinates {(1, 0) (1, 2)};
			\addplot +[mark=none,lightgray,dotted] coordinates {(2, 0) (2, 2)};
			
			
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:1,0.7) {}; 
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:2,1.5) {}; 
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:2,1.1) {}; 
			
			
			\addplot +[mark=none,red,solid,style=thick,<->] coordinates {(2, 1.1) (2, 1.5)};
			\end{axis}
			\end{tikzpicture}
		\end{center}
		\caption{The thick black curve is the graph of a function $f$ and the thinner black line is its tangent line $f(a) + df_a$ are depicted in black. On a small input $h$, the ``error in approximation'' function $r$ outputs the vertical distance indicated in red.}  \label{figure-error-in-approximation}
	\end{figure}
	Rearranging the definition of $r$, we have
	\[ f(a+h) = f(a) + df_a(h) + r(h). \]
	This equation just says that $f(a+h)$ is the sum of the value of the linear approximation (which is $f(a) + df_a(h)$) and the error (which is $r(h)$). 
	
	Notice that $r(0) = 0$. Moreover, since $f$ is differentiable and therefore continuous at $a$ by \cref{differentiable-implies-continuous-single}, we can see that $r$ is continuous at 0. Finally, $r$ is ``small.'' More precisely, we know that $|r(h)| = o(|h|)$ as $h \to 0$ by the discussion above, which means that
	\[ \lim_{h \to 0} \left| \frac{r(h)}{h} \right| = 0 \quad \implies \quad \lim_{h \to 0} \frac{r(h)}{h} = 0. \]
\end{remark} 

As a first example of an application of this idea, here is a weak version of L'H\^opital's rule. 

\begin{exercise}[L'H\^opital's rule, weak version] \label{weak-lhopital}
	Suppose $f, g : S \to \R$ are differentiable at an interior point $a \in S$, that $f(a) = g(a) = 0$, and that $g'(a) \neq 0$. Show that
	\[ \lim_{x \to a} \frac{f(x)}{g(x)} = \frac{f'(a)}{g'(a)}. \]
	\textit{Hint}. If $x = a+h$, we have
	\[ f(x) = f(a+h) = f(a) + df_a(h) + r(h) = f'(a)h + r(h), \]
	where $r$ is the ``error in approximation'' function for $f$ as in \cref{error-in-approximation}. 
\end{exercise}

\section{Rules of differentiation}

The goal in this section is to prove the rules for differentiation that one normally encounters in an introductory calculus course. 

\subsection{Sum and scalar multiples rule}

The following can be proved using either the definition of the derivative \ref{single-var-derivative-definition} or the equivalent ``best linear approximation'' characterization of \cref{unique-linear-single-var}. My suggestion is that you try both methods for these exercises. When you use \cref{unique-linear-single-var}, you may find \cref{little-o-vector-space} useful. 

\begin{exercise}[Constant functions rule] \label{constant-deriv-zero}
	Prove that, if $f : S \to \R$ is constant (ie, there exists a real number $c$ such that $f(x) = c$ for all $x \in S$), then $f'(a) = 0$ for all interior points $a \in S$. 
\end{exercise}

\begin{exercise}[Sum rule] \label{sum-rule-single}
	Prove that, if $f, g : S \to \R$ are both differentiable at an interior point $a \in U$, then $f + g$ is also differentiable at $a$ and \[ (f+g)'(a) = f'(a) + g'(a). \]
\end{exercise}

\begin{exercise}[Scalar multiples rule] \label{scalar-multiples-rule-single}
	Prove that, if $c$ is a constant and $f : S \to \R$ is differentiable at an interior point $a \in S$, then $cf$ is also differentiable at $a$ and \[ (cf)'(a) = c f'(a). \]
\end{exercise}

\subsection{Product and quotient rules}

\begin{proposition}[Product rule] \label{product-rule}
	Suppose $f, g : U \to \R$ are both differentiable at $a \in U$. Then $fg$ is also differentiable at $a$ and 
	\[ (fg)'(a) = f'(a)g(a) + f(a)g'(a). \]
\end{proposition}

\begin{proof}
	The proof is a clever algebraic manipulation of difference quotients. The key trick is introducing a ``cross term'' of the form $f(a)g(a+h)$, and then cancelling it out by also introducing its negative. More precisely, for any sufficiently small real number $h$, we have 
	\[ \begin{aligned} \frac{(fg)(a+h)-(fg)(a)}{h} &= \frac{f(a+h)g(a+h) - f(a)g(a)}{h} \\ 
	&= \frac{f(a+h)g(a+h) {\color{blue}-f(a)g(a+h) + f(a)g(a+h)} - f(a)g(a)}{h} \\
	&= \frac{f(a+h)g(a+h) - f(a)g(a+h)}{h} + \frac{f(a)g(a+h) - f(a)g(a)}{h} \\
	&= \frac{f(a+h) - f(a)}{h} \cdot g(a+h) + f(a) \cdot \frac{g(a+h) - g(a)}{h}. \end{aligned} \]
	Taking the limit as $h \to 0$ yields the result. 
\end{proof}

\begin{exercise}
	Check your understanding of the above proof of the product rule. 
\begin{enumerate}[(a)]
	\item Could you have instead introduced a cross term of the form $f(a+h)g(a)$? If so, rewrite the proof using this cross term. If not, explain why not. 
	\item At what point is \cref{differentiable-implies-continuous-single} tacitly used in the above proof?
\end{enumerate}
\end{exercise}

\begin{exercise}
	 Reprove the scalar multiples rule (\cref{scalar-multiples-rule-single}) using the product rule. 
\end{exercise}

\begin{exercise}[Quotient rule] \label{quotient-rule}
	Suppose $f, g : S \to \R$ are both differentiable at an interior point $a \in U$, and that $g(a) \neq 0$. Prove that $f/g$ is also differentiable at $a$ and 
	\[ (f/g)'(a) = \frac{f'(a)g(a) - f(a)g'(a)}{g(a)^2}. \] 
	\textit{Hint.} Observe that \[ \frac{(f/g)(a+h)-(f/g)(a)}{h} = \frac{\frac{f(a+h)}{g(a+h)} - \frac{f(a)}{g(a)} }{h}. \]
	Multiply both the numerator and denominator by $g(a+h)g(a)$ to clear the denominators inside the numerator, and then introduce the cross term $f(a)g(a)$. Be sure to notice at what point you use \cref{differentiable-implies-continuous-single}.
\end{exercise}

\begin{comment}
	Again, the proof is a clever algebraic manipulation of difference quotients! There are now two tricks: clearing denominators, and then introducing a cross term $f(a)g(a)$. 
	\[ \begin{aligned} \frac{(f/g)(a+h)-(f/g)(a)}{h} &= \frac{\frac{f(a+h)}{g(a+h)} - \frac{f(a)}{g(a)} }{h} \\ 
	&= \frac{g(a+h)g(a)}{g(a+h)g(a)} \cdot \frac{\frac{f(a+h)}{g(a+h)} - \frac{f(a)}{g(a)} }{h} \\
	&= \frac{1}{g(a+h)g(a)} \frac{f(a+h)g(a) - f(a)g(a+h)}{h} \\
	&= \frac{1}{g(a+h)g(a)} \frac{f(a+h)g(a) {\color{blue}- f(a)g(a) + f(a)g(a)} - f(a)g(a+h)}{h} \\ 
	&= \frac{1}{g(a+h)g(a)} \left( \frac{f(a+h)g(a) - f(a)g(a)}{h} + \frac{f(a)g(a) - f(a)g(a+h)}{h} \right) \\
	&= \frac{1}{g(a+h)g(a)} \left( \frac{f(a+h) - f(a)}{h} \cdot g(a) + f(a) \cdot \frac{g(a) - g(a+h)}{h} \right) \end{aligned} \]
\end{comment}

\begin{exercise}[Power rule for integer exponents] \label{power-rule-integer}
	Extend the result of \cref{power-rule-positive} to arbitrary integer exponents. \textit{Hint.} Use the quotient rule for negative integers. The exponent zero is a special but easy case; deal with it separately. 
\end{exercise}

\subsection{Chain rule}

\begin{proposition}[Chain rule] \label{chain-rule-single}
	Suppose that $S$ and $T$ are both subsets of $\R$, that $f : S \to T$ is differentiable at an interior point $a$, that $f(a)$ is an interior point of $T$, and that $g : T \to \R$ is differentiable at $f(a)$. Then the composite $g \circ f : S \to \R$ is also differentiable at $a$, and 
	\[ (g \circ f)'(a) = g'(f(a))f'(a). \]
\end{proposition}

Let us discuss two different (but similar) proofs of this important result. 

\subsubsection{First proof} \label{chain-rule-single-first-proof}

The first proof uses the definition of the derivative \ref{single-var-derivative-definition}, but there is a subtle technicality involved; so, let us discuss it informally first before diving into the formal proof.  

The idea involved is again a clever rewriting of difference quotients. Again, we introduce a cross term, which in this is $f(a+h) - f(a)$. But this time we introduce this cross term ``multiplicatively'' rather than ``additively.'' More precsiely, we have
\[ \begin{aligned} \frac{(g \circ f)(a+h) - (g \circ f)(a)}{h} &= \frac{g(f(a+h)) - g(f(a))}{h} \\ &= \frac{g(f(a+h)) - g(f(a))}{f(a+h)-f(a)} \cdot \frac{f(a+h)-f(a)}{h}. \end{aligned} \]
The difference quotient on the right is precisely $f'(a)$. Thus, it would be sufficient to prove  
\begin{equation} \label{wrong-conjecture} \lim_{h \to 0} \frac{g(f(a+h)) - g(f(a))}{f(a+h)-f(a)} = g'(f(a)). \end{equation}
This formula looks an awful lot like the difference quotient that is used to define the derivative of $g$ at $f(a)$. In fact, since $f$ is continuous at $a$ by \cref{differentiable-implies-continuous-single}, we have $f(a+h) \approx f(a)+h$ for small values of $h$, which means that
\[ \frac{g(f(a+h)) - g(f(a))}{f(a+h)-f(a)} \approx \frac{g(f(a)+h) - g(f(a))}{f(a)+h-f(a)} = \frac{(g \circ f)(a+h) - (g\circ f)(a)}{h} \]
for small values of $h$. Taking the limit as $h \to 0$ on the right-hand side yields precisely $g'(f(a))$, so \cref{wrong-conjecture} seems like a reasonable conjecture. 

The problem is that \cref{wrong-conjecture} isn't true. Issues arise when $f(a+h) = f(a)$ for arbitrarily small values of $h$, which leads to an infinite sequence of divisions by 0. 

\begin{example}
	Let \[ f(x) = \begin{cases} x^2\sin(1/x) & \text{if } x \neq 0 \\ 0 & \text{if } x = 0 \end{cases} \] and let $g(x) = x^2$. Observe that $g'(f(0)) = 0$. You will show later (in \cref{power-times-sine}) that $f$ is in fact differentiable at 0, so these functions do satisfy the hypothesis of the chain rule. However, the limit as $h \to 0$ of
	\[ \frac{g(f(a+h)) - g(f(a))}{f(a+h)-f(a)} = \frac{(h^2\sin(1/h))^2}{h^2\sin(1/h)} \]
	does not exist. For example, the sequence $h_k = 1/k\pi$ converges to 0 as $k \to \infty$, but the expression above is undefined along this sequence. 
\end{example}

We will get around the kind of issue that is described by the above example by carefully studying the expression
\begin{equation} \label{weird-difference-quotient} \frac{g(f(a+h)) - g(f(a))}{f(a+h)-f(a)}. \end{equation}
Observe that, when $f(a+h)\neq f(a)$, then this expression is precisely the difference quotient that calculates the slope of the secant line passing through $(f(a), g(f(a))$ and $(f(a+h), g(f(a+h))$. When $f(a+h) = f(a)$, this secant line doesn't make sense; but what does make sense is the tangent line! So we'll remove the undefinedness of expression \eqref{weird-difference-quotient} by extending it so that it takes the value $g'(f(a))$ whenever $f(a+h) = f(a)$. 

This leads us to our first proof of the chain rule. 

\begin{proof}[First proof of the chain rule]
	Consider the function $\epsilon$ defined for small values of $h$ by $\epsilon(h) = f(a+h)-f(a)$. See \cref{value-difference-chain-rule}. You will verify in \cref{chain-rule-details} below that $\epsilon$ is continuous at 0. 
	
	\begin{figure}
		\begin{center}
			\begin{tikzpicture}
			\begin{axis}[
			xmin=-0.5,xmax=3,
			ymin=-0.5,ymax=2.25,
			ytick={0.7, 1.5},
			xtick={1,2},
			xticklabels={$a$, $a+h$},
			yticklabels={$f(a)$,$f(a+h)$},
			axis lines=middle,
			axis line style=lightgray,
			width=0.75\textwidth
			]
			\addplot[black,style=thick,<->] expression[domain=-0.5:2.45,samples=50]{(x*(x^2+1)+5)/10}; 
			
			
			\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{0.7};
			\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{1.5}; 
			\addplot +[mark=none,lightgray,dotted] coordinates {(1, 0) (1, 2)};
			\addplot +[mark=none,lightgray,dotted] coordinates {(2, 0) (2, 2)};
			
			
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:1,0.7) {}; 
			
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:2,1.5) {}; 
			
			
			\addplot +[mark=none,gray,red,<->,solid] coordinates {(2, 0.7) (2, 1.5)};
			\end{axis}
			\end{tikzpicture}
		\end{center}
		\caption{The graph of $f$ is depicted in black. The value of the function $t$ on a small nonzero input value $h$ is the vertical distance depicted in red.}  \label{value-difference-chain-rule}
	\end{figure}

	Next, define a function $\sigma$ by the following.  
	\[ \sigma(k) = \begin{cases} \dfrac{g(f(a)+k) - g(f(a))}{k} & \text{if } k \neq 0 \\ g'(f(a)) & \text{if } k = 0 \end{cases} \]
	In other words, $\sigma$ is the function that, on some small nonzero input $k$, outputs the slope of the secant line connecting the two points on the graph of $g$ where $x = f(a)$ and $x = f(a) + k$. See \cref{slope-function-chain-rule}. It follows from the definition of the derivative of $g$ at $f(a)$ that the function $\sigma$ is continuous at 0. 
	
	\begin{figure}
		\begin{center}
			\begin{tikzpicture}
			\begin{axis}[
			xmin=-0.5,xmax=3,
			ymin=-0.5,ymax=2.25,
			ytick={1, 2},
			xtick={1,2},
			xticklabels={$f(a)$, $f(a)+k$},
			yticklabels={$g(f(a))$,$g(f(a)+k)$},
			axis lines=middle,
			axis line style=lightgray,
			width=0.75\textwidth
			]
			\addplot[black,style=thick,<->] expression[domain=0.25:2.02,samples=100]{4*x^3-12*x^2+9*x}; 
			
			
			\addplot[red,<->] expression[domain=0.2:2.2,samples=2]{(x-1)+1}; 
			
			
			\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{1};
			\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{2}; 
			\addplot +[mark=none,lightgray,dotted] coordinates {(1, 0) (1, 2)};
			\addplot +[mark=none,lightgray,dotted] coordinates {(2, 0) (2, 2)};
			
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:1,1) {}; 
			
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:2,2) {}; 
			\end{axis}
			\end{tikzpicture}
		\end{center}
		\caption{The graph of $g$ is depicted in black. The output of the function $\sigma$ on a small nonzero input value $k$ is the slope of the red secant line passing through the two points $(f(a), g(f(a))$ and $(f(a)+k, g(f(a)+k))$. }  \label{slope-function-chain-rule}
	\end{figure}
	
	
	Observe that
	\begin{equation} \label{alternative-to-multiplicative-cross-term} \frac{(g \circ f)(a+h) - (g \circ f)(a)}{h} = \sigma(\epsilon(h)) \cdot \frac{f(a+h)-f(a)}{h} \end{equation}
	for all small values of $h$, as you will verify in \cref{chain-rule-details} below. Thus
	\[ \begin{aligned} \lim_{h \to 0} \frac{(g \circ f)(a+h) - (g \circ f)(a)}{h} &= \lim_{h \to 0} \sigma (\epsilon(h)) \cdot \frac{f(a+h)-f(a)}{h} \\ 
	&\stackrel{(*)}{=} \sigma (\epsilon(0)) f'(a) \\
	&= g'(f(a)) f'(a) \end{aligned}  \]
	where we have used the continuity of $\epsilon$ and $\sigma$ and the definition of the derivative of $f$ at $a$ for the step labeled $(*)$ above. This proves the chain rule. 
\end{proof}
	
\begin{exercise} \label{chain-rule-single-first-proof-details}
	Verify the missing details in the above proof. Specifically, check the following. 
	\begin{enumerate}[(a)]
		\item Prove that $\epsilon$ is continuous at 0.
		\item Prove \cref{alternative-to-multiplicative-cross-term}. \textit{Hint.} As suggested by our analysis before the proof of the chain rule, split up your proof of \cref{alternative-to-multiplicative-cross-term} into two cases: when $f(a+h) \neq f(a)$, and when $f(a+h) = f(a)$. 
	\end{enumerate}
\end{exercise}

\subsubsection{Second proof} \label{chain-rule-single-second-proof}

Our second proof of the chain rule will use the ``best linear approximation'' interpretation of derivatives provided by \cref{unique-linear-single-var}. 

\begin{proof}[Second proof of the chain rule] \renewcommand{\qedsymbol}{}
	By \cref{unique-linear-single-var}, it is sufficient to prove that
	\[ g(f(a+h))-g(f(a))-dg_{f(a)}(df_a(h)) = o(|h|) \text{ as } h \to 0. \]
	Define the following ``error in approximation'' functions, as in \cref{error-in-approximation}. 
	\[ \begin{aligned} 
	r(h) &= f(a+h)-f(a) - df_a(h) \\ 
	s(k) &= g(f(a)+k)-g(f(a))-dg_{f(a)}(k) \end{aligned} \]
	Since
	\[ f(a+h) = f(a) + df_a(h) + r(h) \]
	by definition of $r$, we have
	\[ \begin{aligned}
	g(f(a+h)) &= g(f(a) + df_a(h) + r(h)) \\
	&= g(f(a)) + dg_{f(a)}(df_{a}(h) + r(h)) + s(df_a(h)+r(h)) \\
	&= g(f(a)) + dg_{f(a)}(df_{a}(h)) + dg_{f(a)}(r(h)) +  s(df_a(h) + r(h)) \end{aligned} \]
	where we use the definition of $s$ on input $k = df_a(h)+r(h)$ for the second step, and the fact that $dg_{f(a)}$ is linear for the third. Thus
	\[ g(f(a+h)) - g(f(a)) - dg_{f(a)}(df_a(h)) = dg_{f(a)}(r(h)) + s(df_a(h)+r(h)). \]
	We therefore want to prove that the right-hand side is $o(|h|)$. 	
	
	Since \[ dg_{f(a)}(r(h)) = g'(f(a)) \cdot r(h) \] is just a scalar multiple of $r$, and $r = o(|h|)$, we see that $dg_{f(a)}(r) = o(|h|)$ also. So it is sufficient to prove that
	\[ s(df_a(h)+r(h)) = o(|h|). \]
	We've now hit the technical part, so let's hit pause. 
\end{proof}

To see why this is technical, and for essentially the same reasons as the previous proof, notice that we're trying to prove that
\[ \lim_{h \to 0} \frac{|s(df_a(h)+r(h)|)}{|h|} = 0.  \]
To prove this, it's tempting to introduce a multiplicative cross-term as follows. 
\[ \frac{|s(df_a(h)+r(h)|)}{|h|} = \frac{|s(df_a(h)+r(h)|)}{|df_a(h)+r(h)|} \cdot \frac{|df_a(h)+r(h)|}{|h|} \]
But notice that $df_a(h)+r(h) = f(a+h)-f(a)$ by definition of $r$, so this cross term is problematic for the same reasons as before! If $f(a+h) = f(a)$ for $h$ arbitrary close to $0$, we have an infinite sequence of divisions by zeroes that causes problems. 

We'll deal with this problem in essentially the same way as before: namely, we extend the definition of 
\[\frac{|s(df_a(h)+r(h)|)}{|df_a(h)+r(h)|}\]
so that it is also defined when $f(a+h) = f(a)$. 

\begin{proof}[Second proof of the chain rule, continued]
	Define $\eta(k) = s(k)/k$ for nonzero $k$. If we set $\eta(0) = 0$, the fact that $s(k) = o(|k|)$ implies that $\eta$ is continuous at 0. Then 
	\[ \begin{aligned} \frac{|s(df_a(h)+r(h))|}{|h|} &= \frac{|\eta(df_a(h)+r(h))| \cdot |df_a(h)+r(h)|}{|h|} \\
	&\leq |\eta(df_a(h)+r(h))| \cdot \left( |f'(a)| + \frac{|r(h)|}{|h|} \right). \end{aligned} \]
	Observe that $r$ and therefore $df_a + r$ is continuous at $0$ (cf. \cref{error-in-approximation}). Moreover, we have just seen that $\eta$ is continuous at $(df_a + r)(0) = 0$. Thus the composite $\eta \circ (df_a + r)$ is also continuous at 0. Thus
	\[ \lim_{h \to 0} |\eta(df_a(h)+r(h))| \cdot \left( |f'(a)| + \frac{|r(h)|}{|h|} \right) = 0, \]
	where we have used the fact that $|r(h)| = o(|h|)$ to see that the parenthetical expression has a finite limit (namely, $|f'(a)|$). So, by the squeeze theorem, we can therefore conclude that $|s(df_a(h)+r(h))| = o(|h|)$.  
\end{proof}

\begin{comment}[Second proof of the chain rule, continued]
	To prove that \[ s(df_a(h)+r(h)) = o(|h|), \] let's whip out some epsilons and deltas. Suppose $\epsilon > 0$. We want to find a $\delta > 0$ such that $|h| < \delta$ implies that
	\[ |s(df_a(h)+r(h))| \leq \epsilon|h|. \]
	
	Since $s$ is $o(|h|)$, we know that for any $\epsilon_1 > 0$ there exists $\delta_1 > 0$ such that $|k| < \delta_1$ implies that
	\[ \begin{aligned} |s(k)| \leq \epsilon_1|k|. \end{aligned} \] 
	Since $df_a + r$ is continuous at 0, there then exists $\delta_2 > 0$ such that $|h| < \delta$ implies that $|df_a(h) + r(h)| < \delta_1$. Then for $|h| < \delta_2$, we have
	\[ |s(df_a(h) + r(h)| \leq \epsilon_1 |df_a(h) + r(h)| 
	\leq \epsilon_1|f'(a)||h| + \epsilon_1|r(h)| \]
	Since $r$ is $o(|h|)$, for any $\epsilon_2 > 0$ there exists $\delta_3 > 0$ such that $|h| < \delta_3$ implies that $|r(h)| \leq \epsilon_2 |h|$. Let $\delta = \min\{\delta_2, \delta_3\}$. Then for $|h| < \delta$, we can continue the above chain of inequalities to obtain 
	\[ |s(df_a(h) + r(h)| \leq \epsilon_1|f'(a)||h| + \epsilon_1 \epsilon_2|h| = \epsilon_1(|f'(a)| + \epsilon_2)|h|. \]
	So we let $\epsilon_2 = 1$ and then let 
	\[ \epsilon_1 = \frac{\epsilon}{|f'(a)| + 1}. \]
	The $\delta$ obtained by going through the above process then has the desired property. 
\end{comment}

Notice that in the first proof, the trick to overcoming the technicality was defining the function $\sigma$ at $k = 0$. In the second proof, the trick to overcoming the same technicality was defining $\eta$ at $k = 0$. These two tricks are really the same trick, as shown by the following.  

\begin{exercise}
	Prove that the function $\eta$ from the second proof and the function $\sigma$ from the first proof are related by the equation
	\[ \sigma(k) = g'(f(a)) + \eta(k). \]
\end{exercise}

\subsubsection{Applications of the chain rule}

\begin{exercise}[Derivatives of inverses] \label{inverse-derivative}
	Suppose $S$ and $T$ are both subsets of $\R$ and $f : S \to T$ is bijective, with inverse function $f^{-1} : T \to S$. Suppose further that $f$ is differentiable at an interior point $a \in S$, that $f'(a) \neq 0$, and that $b = f(a)$ is an interior point of $T$. Prove that $f^{-1}$ is differentiable at $b$ and
	\[ (f^{-1})'(b) = \frac{1}{f'(a)} = \frac{1}{f'(f^{-1}(b))}. \]
\end{exercise}

\begin{exercise}[Power rule for fractional exponents] \label{power-rule-fractional}
	Extend the result of \cref{power-rule-integer} to arbitrary rational exponents (ie, to exponents that can be written in the form $m/n$ where $m$ and $n$ are both integers). 
\end{exercise}

\subsection{Some rules without proofs}

In this section, we discuss differentiation of some common ``transcendental functions.'' First off, here is the most general statement of the power rule, but you should note that \cref{power-rule-fractional} already covers the most common and most important cases. 

\begin{proposition}[Power rule for real exponents] \label{power-rule-general}
	If $r$ is any real number and $f : [0, \infty) \to \R$ is the function $f(x) = x^r$, then $f$ is differentiable and $f'(x) = rx^{r-1}$. 
\end{proposition}

The first step in proving this result is giving a rigorous construction of the real numbers and then rigorously defining what $x^r$ even means for non-rational exponents $r$. This would be a long detour, so it is omitted. In fact, once we've done this, it turns out the power rule for arbitrary real exponents is actually a relatively straightforward \emph{consequence} of \cref{power-rule-fractional}. 

Next up, here is a fact about differentiating power series. The proof is omitted because it is rather technical. 

\begin{proposition} \label{differentiating-power-series}
	Suppose the power series $\sum_{n = 0}^\infty a_n x^n$
	has radius of convergence $R$. Then the function $f : (-R, R) \to \R$ given by
	\[ f(x) = \sum_{n = 0}^\infty a_n x^n \]
	is differentiable and
	\[ f'(x) = \sum_{n = 0}^\infty na_n x^{n-1}. \]
	Moreover, the radius of convergence of the power series $\sum_{n = 0}^\infty na_n x^{n-1}$
	is also $R$. 
\end{proposition}

The exponential, sine and cosine functions are usually defined using power series. More precisely, we have the following. 
\[ \begin{aligned} \exp(x) &= \sum_{n = 0}^\infty \frac{x^n}{n!} \\
\cos(x) &= \sum_{n = 0}^\infty \frac{(-1)^n x^{2n}}{(2n)!} \\
\sin(x) &= \sum_{n = 0}^\infty \frac{(-1)^n x^{2n+1}}{(2n+1)!} \end{aligned} \]
All three of these power series have infinite radius of convergence, so the following three facts are immediate consequences of \cref{differentiating-power-series}. 
\begin{enumerate}[(1)]
	\item If $f(x) = \exp(x)$, then $f'(x) = \exp(x)$. 
	\item If $f(x) = \sin(x)$, then $f'(x) = \cos(x)$. 
	\item If $f(x) = \cos(x)$, then $f'(x) = -\sin(x)$.
\end{enumerate} 

\begin{remark}
	Defining the exponential, sine, and cosine functions using power series is mathematically fine, but there are numerous facts about these functions that are not clear from the power series, and one needs to prove these facts separately. For example, it's useful knowing that, if $e := \exp(1)$, then $\exp(x) = e^x$. This fact which is not immediately clear from the power series definition and needs to be proved; we won't prove it here, but you should feel free to use it. It's also useful to know that sines and cosines have something to do with trigonometry. Again, we won't prove these facts, but you should feel free to use them. 
\end{remark}

From this, we can derive the derivative of the logarithm.

\begin{exercise}
	If $f(x) = \ln(x)$, show that $f'(x) = 1/x$. 
	\textit{Hint}. The natural logarithm $\ln : (0, \infty) \to \R$ is the inverse of the exponential function. 
\end{exercise}

Here is some further practice with differentiation of transcendental functions. 

\begin{exercise} \label{power-times-sine}
	\begin{enumerate}[(a)]
		\item Prove that the function \[ f(x) = \begin{cases} x\sin(1/x) & \text{if } x \neq 0 \\ 0 & \text{if } x = 0 \end{cases} \]
		is continuous, but that it is not differentiable at 0. 
		\item Prove that the function \[ f(x) = \begin{cases} x^2\sin(1/x) & \text{if } x \neq 0 \\ 0 & \text{if } x = 0 \end{cases} \]
		is differentiable, but that the derivative $f'$ is not continuous at 0.  
	\end{enumerate} 
\end{exercise}

\section{Derivatives, extremums, and monotonicity}

\subsection{Derivatives and local extremums}

We first recall the following definition. 

\begin{definition}
	If $X$ is a metric space and $f : X \to \R$ is a function, a point $a \in X$ is said to be a \emph{local maximum} of $f$ if there exists a neighborhood $U$ of $a$ in $X$ such that $f(x) \leq f(a)$ for all $x \in U$. We define \emph{local minimum} by reversing the inequality. A \emph{local extremum} is a point that is either a local maximum or a local minimum.
\end{definition}

The following is a key property of derivatives that you probably remember from calculus. 

\begin{exercise}[Interior extremum theorem] \label{interior-extremum}
	Suppose an interior point $a \in S$ is a local extremum of a function $f : S \to \R$. If $f$ is differentiable at $a$, prove that $f'(a) = 0$. \textit{Hint}. Consider the following function (cf. the first proof of the chain rule in \cref{chain-rule-single-first-proof}).
	\[ \sigma(h) = \begin{cases} \dfrac{f(a+h)-f(a)}{h} & \text{if } h \neq 0 \\ f'(a) & \text{if } h = 0 \end{cases} \]
	Since $f$ is differentiable at $x = a$, we know that $\sigma$ is continuous at $h = 0$. If $a$ is a local minimum, what can you say about the values of $\sigma$ for nonzero values of $h$? What about if $a$ is a local maximum?
\end{exercise}

But the reader is reminded that the converse to the interior extremum theorem \ref{interior-extremum} is not true. 

\begin{exercise} \label{interior-extremum-converse-false}
	Give an example of a function $f : \R to \R$ where $f'(0) = 0$, but $f$ does not have a local extremum at 0. 
\end{exercise}

\subsection{Mean value theorem}

The mean value theorem \ref{mean-value-theorem} is an extremely important foundational result; it comes up often during proofs. Rolle's theorem \ref{rolles-theorem} is a special case of the mean value theorem, but the general case can be derived from this special case. 

\begin{exercise}[Rolle's theorem] \label{rolles-theorem}
	Suppose $a < b$ and let $f : [a, b] \to \R$ be a differentiable function. If $f(a) = f(b) = 0$, prove that there exists $c \in (a, b)$ such that $f'(c) = 0$. \textit{Hint}. Any function must fall into at least one of the following three cases: (i) $f$ is constantly zero, (ii) $f$ takes on positive values somewhere on $[a, b]$, and (iii) $f$ takes on negative values somewhere on $[a, b]$. Deal with the first case using \cref{constant-deriv-zero}, and the latter two cases using the extreme value theorem\footnotemark\ together with \cref{interior-extremum}. 
\end{exercise}

\footnotetext{The \emph{extreme value theorem} says that if $K$ is a compact metric space and $f : K \to \R$ is continuous, then $f$ achieves its global maximum and its global minimum somewhere on $K$ (cf. \cite[theorems 3.17 and 6.30]{protter-morrey}).}

\begin{exercise}[Mean value theorem] \label{mean-value-theorem}
	Suppose $a < b$ and let $f : [a, b] \to \R$ be a differentiable function. Prove that there exists $c \in (a, b)$ such that
	\[ f'(c) = \frac{f(b) - f(a)}{b-a}. \]
	\textit{Hint}. Let $\ell$ be the secant line passing through $(a, f(a))$ and $(b, f(b))$. Then consider the function $g = f - \ell$. See \cref{subtracting-the-secant}. 
	\begin{figure}
		\begin{center}
			\begin{tikzpicture}
			\begin{axis}[
			xmin=-0.1,xmax=2.5,
			ymin=-0.1,ymax=2,
			ytick={0.5, 1.5},
			xtick={0.5,2},
			xticklabels={$a$, $b$},
			yticklabels={$f(a)$, $f(b)$},
			axis lines=middle,
			axis line style=lightgray,
			width=0.5\textwidth
			]
			\addplot[black,thick,-] expression[domain=0.5:2,samples=50]{2/3*x+1/6 + 2*(x-0.5)*(x-1.5)*(x-2)}; 
			
			
			\addplot[red,<->] expression[domain=0.25:2.25,samples=2]{2/3*x+1/6}; 
			
			
			\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{0.5};
			\addplot[lightgray,dotted] expression[domain=0:2.75,samples=2]{1.5}; 
			\addplot +[mark=none,lightgray,dotted] coordinates {(0.5, 0) (0.5, 2)};
			\addplot +[mark=none,lightgray,dotted] coordinates {(2, 0) (2, 2)};
			
			
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:0.5,0.5) {}; 
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:2,1.5) {}; 
			\end{axis}
			\end{tikzpicture}\hspace{1cm}
			\begin{tikzpicture}
			\begin{axis}[
			xmin=-0.1,xmax=2.5,
			ymin=-1,ymax=1.1,
			ytick={0},
			xtick={0.5,2},
			xticklabels={$a$, $b$},
			yticklabels={},
			axis lines=middle,
			axis line style=lightgray,
			width=0.5\textwidth
			]
			\addplot[black,thick,-] expression[domain=0.5:2,samples=50]{2*(x-0.5)*(x-1.5)*(x-2)}; 
			
			\addplot[red,<->] expression[domain=0.25:2.25,samples=2]{0};
			
			\addplot +[mark=none,lightgray,dotted] coordinates {(0.5, -2) (0.5, 2)};
			\addplot +[mark=none,lightgray,dotted] coordinates {(2, -2) (2, 2)};
			
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:0.5,0) {}; 
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:2,0) {}; 
			\end{axis}
			\end{tikzpicture}
		\end{center}
		\caption{On the left is depicted the graph of a function $f$ and the secant line $\ell$. Subtracting $\ell$ from the picture on the left yields the picture on the right. }  \label{subtracting-the-secant}
	\end{figure}
	Then $g(a) = g(b) = 0$, so we can apply Rolle's theorem \ref{rolles-theorem} to $g$. 
\end{exercise}

\subsection{Derivatives and monotonicity}

Here is a standard property of derivatives from calculus; the proof follows quickly from the mean value theorem. 

\begin{lemma} \label{deriv-zero-implies-constant}
	Suppose $I$ is an open interval and that $f : I \to \R$ is differentiable. If $f' = 0$, then $f$ is constant. 
\end{lemma}

\begin{proof}
	Suppose we have $a < b$ are two distinct points in $I$. Since $f$ is differentiable on $I$ and $[a,b] \subseteq I$, we know in particular that $f$ is also differentiable on $[a,b]$, so the mean value theorem applies: there exists a point $c \in (a,b)$ such that
	\[ \frac{f(b)-f(a)}{b-a} = f'(c). \]
	But $f'(c) = 0$ by assumption, so $f(a) = f(b)$. 
\end{proof}

Next, we recall the following definition. 

\begin{definition}
	Suppose $f : S \to \R$ is a function and $x, y \in S$. 
	\begin{itemize}
		\item $f$ is \emph{increasing} if $x < y$ implies $f(x) \leq f(y)$. 
		\item $f$ is \emph{decreasing} if $x < y$ implies $f(x) \geq f(y)$.
		\item $f$ is \emph{strictly increasing} if $x < y$ implies $f(x) < f(y)$. 
		\item $f$ is \emph{strictly decreasing} if $x < y$ implies $f(x) > f(y)$. 
	\end{itemize}
Further, $f$ is \emph{monotonic} if it is either increasing or decreasing, and \emph{strictly monotonic} if it is either strictly increasing or strictly decreasing. 
\end{definition}

The following standard property of differentiable functions again follows quickly from the mean value theorem. 

\begin{exercise} \label{deriv-increasing-decreasing}
	Suppose $I$ is an open interval and that $f : I \to \R$ is differentiable. Prove the following four statements. 
	\begin{enumerate}[(a)]
		\item If $f' \geq 0$, then $f$ is increasing. 
		\item If $f' > 0$, then $f$ is strictly increasing. 
		\item If $f' \leq 0$, then $f$ is decreasing. 
		\item If $f' < 0$, then $f$ is strictly decreasing. 
	\end{enumerate} 
	\textit{Hint}. For all four, imitate the proof of \cref{deriv-zero-implies-constant}. 
\end{exercise}

\subsection{Darboux's theorem $\ast$}

Darboux's theorem is not commonly discussed in elementary calculus, nor is it strictly necessary for what follows. But it is an interesting fact about derivatives, and is sometimes useful to know. Before getting to the statement of the theorem, it is convenient to make the following definition. 

\begin{definition}
	Suppose $I$ is an interval and $f : I \to \R$ is a function. Then $f$ \emph{has the intermediate value property} if, for every $a < b$ in $I$ and every $y \in \R$ strictly between $f(a)$ and $f(b)$, there exists $x \in (a, b)$ such that $f(x) = y$. 
\end{definition}

The intermediate value theorem asserts that every continuous function has the intermediate value property. Simple examples of discontinuous functions don't have the intermediate value property (cf. \cref{removable-jump-ivp} below). But, it turns out that some discontinuous functions do have the intermediate value property (cf. \cref{discontinuous-ivp} below). 

\begin{exercise} \label{removable-jump-ivp}
	Let $I$ be an interval and $f : I \to \R$ a function. 
	\begin{enumerate}[(a)]
		\item Recall that $f$ has a \emph{removable discontinuity} at a point $a \in I$ if
		\[ \lim_{x \to a} f(x) \]
		exists, but the value of this limit does not equal $f(a)$. Show that, if $f$ has a removable discontinuity, then $f$ does \emph{not} have the intermediate value property. 
		 
		\item Recall that $f$ has a \emph{jump discontinuity} at a point $a \in I$ if the two one-sided limits
		\[ \lim_{x \to a^-} f(x) \text{ and } \lim_{x \to a^+} f(x) \]
		both exist, but their values are not equal to each other. Show that, if $f$ has a jump discontinuity, then $f$ does \emph{not} have the intermediate value property.  
	\end{enumerate}
\end{exercise}

\begin{exercise} \label{discontinuous-ivp}
	Show that the function $f : \R \to \R$ defined by
	\[ f(x) = \begin{cases} \sin(1/x) & \text{if } x \neq 0 \\ 0 & \text{if } x = 0 \end{cases} \]
	is discontinuous at $x = 0$ but has the intermediate value property. \textit{Hint}. First show that for any non-degenerate interval $I$ containing 0, we have $f(I) = [-1,1]$. Then explain how this implies the intermediate value property. 
\end{exercise}

We know that the derivative $f'$ of a differentiable function $f$ need not be continuous (cf. \cref{power-times-sine}). Darboux's theorem asserts that, even though $f'$ might not be continuous, it must still have the intermediate value property! This means, for example, that derivatives cannot have removable or jump discontinuities (cf. \cref{removable-jump-ivp}). 

\begin{theorem}[Darboux] \label{darboux}
	Suppose $I$ is an open interval and $f : I \to \R$ is differentiable. Then $f'$ has the intermediate value property. 
\end{theorem}

\begin{proof}
	Suppose $a < b$ are two elements of $I$ and $m$ is a real number strictly between $f'(a)$ and $f'(b)$. We want to show that there exists $c \in (a,b)$ such that $f'(c) = m$. This statement somewhat resembles the statement of the mean value theorem \ref{mean-value-theorem}, and indeed, our proof will also somewhat resemble the proof we gave of the mean value theorem. 
	
	Consider the differentiable function $g : I \to \R$ defined by $g(x) = f(x) - mx$. See \cref{subtracting-a-line}. Observe that $f'(c) = m$ if and only if $g'(c) = 0$. Thus we would like to show that there exists $c \in (a, b)$ such that $g'(c) = 0$. 
	
	
	\begin{figure}
		\begin{center}
			\begin{tikzpicture}
			\begin{axis}[
			xmin=-0.5,xmax=2.5,
			ymin=-0.25,ymax=4,
			xtick={0.5,2},
			ytick={0},
			xticklabels={$a$, $b$},
			yticklabels={},
			axis lines=middle,
			axis line style=lightgray,
			width=0.5\textwidth
			]
			\addplot[black,thick,<->] expression[domain=0.25:2.25,samples=50]{(x-1)^2+1 + x/2}; 
			
			
			\addplot[red,<->] expression[domain=-0.25:2.25,samples=2]{x/2}; 
			
			
			\addplot[lightgray,dashed,<->] expression[domain=-0.25:1,samples=2]{-0.5*(x-0.5) + 1.5};
			\addplot[lightgray,dashed,<->] expression[domain=1.5:2.4,samples=2]{2.5*(x-2) + 3}; 
			
			\addplot +[mark=none,lightgray,dotted] coordinates {(0.5, 0) (0.5, 4)};
			\addplot +[mark=none,lightgray,dotted] coordinates {(2, 0) (2, 4)};
			
			
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:0.5,1.25+0.25) {}; 
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:2,2+1) {}; 
			\end{axis}
			\end{tikzpicture}\hspace{1cm}
			\begin{tikzpicture}
			\begin{axis}[
			xmin=-0.5,xmax=2.5,
			ymin=-1.25,ymax=3,
			ytick={0},
			xtick={0.5,2},
			xticklabels={$a$, $b$},
			yticklabels={},
			axis lines=middle,
			axis line style=lightgray,
			width=0.5\textwidth
			]
			\addplot[black,thick,<->] expression[domain=0.25:2.25,samples=50]{(x-1)^2+1}; 
			
			\addplot[red,<->] expression[domain=-0.25:2.25,samples=2]{0};
			
			\addplot +[mark=none,lightgray,dotted] coordinates {(0.5, -2) (0.5, 3)};
			\addplot +[mark=none,lightgray,dotted] coordinates {(2, -2) (2, 3)};
			
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:0.5,1.25) {}; 
			\node[color=black,draw=black,circle,fill,inner sep=2pt] at (axis cs:2,2) {}; 
			\end{axis}
			\end{tikzpicture}
		\end{center}
		\caption{On the left is depicted the graph of a function $f$, the two tangent lines at $a$ and $b$, and the graph of the line $\ell(x) = mx$ for some $m$ in between $f'(a)$ and $f'(b)$. Subtracting $\ell$ from this picture yields the picture on the right.}  \label{subtracting-a-line}
	\end{figure}
	
	There are two cases two consider: either (i) $f'(a) < m < f'(b)$ or (ii) $f'(a) > m > f'(b)$. Suppose we are in case (i). By the extreme value theorem, $g$ attains its minimum on the compact interval $[a,b]$ at some point $c \in [a,b]$. We claim that $c$ must be an interior point of $[a,b]$, and then interior extremum theorem \ref{interior-extremum} will imply that $g'(c) = 0$. 
	
	If $c = a$ were the minimum of $g$ on $[a,b]$, then we would have 
	\[ \frac{g(a+h)-g(a)}{h} > 0 \]
	for all sufficiently small positive values of $h$. This would mean that $g'(a) \geq 0$, but we know that $g'(a) = f'(a) - m < 0$, which is a contradiction. Similarly, if $c = b$ were the minimum of $g$ on $[a,b]$, then we would have
	\[ \frac{g(b+h)-g(b)}{h} < 0 \]
	for all sufficiently small negative values of $h$. This would imply that $g'(b) \leq 0$, but we know that $g'(b) = f'(b) - m > 0$. 
	
	In case (ii), we instead let $c$ be the \emph{maximum} of $g$ on $[a,b]$, and argue similarly to show that $c$ must be an interior point. 
\end{proof}

\section{Levels of differentiability} 

\subsection{$C^k$ functions}

\begin{definition}
	Suppose $U$ is an open subset of $\R$ and $f : U \to \R$ is differentiable. If the derivative $f' : U \to \R$ is also continuous, then $f$ is said to be \emph{continuously differentiable}. 
\end{definition}

\begin{definition} \label{cn-single-var}
	Let $U$ be an open subset of $\R$ and $f : U \to \R$ a function. We make the following string of definitions. 
	\begin{itemize}
		\item $f$ is $C^0$ if it is continuous. 
		\item $f$ is $C^1$ if it is continuously differentiable. 
		\item $f$ is $C^2$ if the derivative $f'$ is $C^1$. In other words, this means that the derivative of the derivative $(f')'$, denoted $f''$ or $f^{(2)}$ and called the \emph{second derivative}, exists and is continuous. 
		\item In general, for any positive integer $k$, we say that $f$ is $C^k$ if $f'$ exists and is $C^{k-1}$. This means that the \emph{$k$th derivative} $f^{(k)} = (f^{(k-1)})'$ exists and is continuous.
	\end{itemize}
\end{definition}

\begin{exercise} \label{sums-scalars-cn-singlevar}
	Prove that the set of all $C^k$ functions $U \to \R$ is a vector space. 
\end{exercise}

\begin{exercise}[General Leibniz Rule] \label{product-cn-singlevar}
	Suppose that $f, g : U \to \R$ are $C^k$ functions. Prove that the product $fg$ is also $C^k$, and that
	\[ (fg)^{(k)} = \sum_{i = 0}^k {k \choose i} f^{(i)} g^{(k-i)}. \]
\end{exercise}

\begin{exercise} \label{composite-cn-singlevar}
	Show that the composite of two $C^k$ functions is also $C^k$. \textit{Hint}. Induct on $k$. The base case $k = 0$ is clear. For the inductive step, use the chain rule \ref{chain-rule-single} and the first part of \cref{product-cn-singlevar}. 
\end{exercise}

Observe that we have an infinite chain of implications
\[ \dotsb \implies C^3 \implies C^2 \implies C^1 \implies C^0. \]
The following exercise shows that all of these implications are strict; in other words, for every $k$, there exists a $C^k$ function which is not $C^{k+1}$.

\begin{exercise}
	For every non-negative integer $k$, let 
	\[ f_k(x) = x^k|x|. \]
	Prove that $f_k$ is $C^k$ but not $C^{k+1}$. \textit{Hint.} First prove that $f_{k+1}' = (k+2)f_k$. 
\end{exercise}

\subsection{Smooth functions} \label{smooth-single-var}

\begin{definition}
	A function $f : U \to \R$ is said to be $C^\infty$, or \emph{infinitely differentiable}, or \emph{smooth}, if it is in $C^k$ for all $k$. 
\end{definition}

It's a little annoying that there are three different words that all mean the same thing, but that's just how it is; all three are commonly used in the mathematical literature, so it's best to get used to all of them. We'll usually use the word ``smooth.'' It follows immediately from \cref{sums-scalars-cn-singlevar,product-cn-singlevar,composite-cn-singlevar} that sums, scalar multiples, products, and composites of smooth functions are also smooth. 

An important principle is that we can tailor smooth functions to almost arbitrary specifications. Our first example of this principle is the following smooth bijection $(-1,1) \to \R$ whose inverse is also smooth. A smooth bijection with a smooth inverse is also called a ``diffeomorphism.''

\begin{example} \label{interval-to-line}
	Consider the function $f : (-1,1) \to \R$ defined by \[ f(x) = \frac{x}{1-x^2}. \]
	See \cref{interval-to-line-graph}. 
	\begin{figure}
		\begin{center}
			\begin{tikzpicture}
			\begin{axis}[
			xmin=-1.5,xmax=1.5,
			ymin=-5,ymax=5,
			ytick={0},
			xtick={-1,1},
			xticklabels={$-1$,1},
			yticklabels={},
			axis lines=middle,
			axis line style=lightgray,
			width=0.75\textwidth
			]
			
			
			\addplot[black,style=thick,<->] expression[domain=-0.9:0.9,samples=100]{x/(1-x^2)};
			\end{axis}
			\end{tikzpicture}
		\end{center}
		\caption{The graph of the function $f$ from \cref{interval-to-line}.}  \label{interval-to-line-graph}
	\end{figure}
	Since this is a quotient of two smooth functions and the denominator is nonzero on $(-1,1)$, this function is also smooth. We can calculate that
	\[ f'(x) = \frac{1+x^2}{(1-x^2)^2} \]
	and we can see from this formula that the derivative is always strictly positive. Thus $f$ is strictly increasing (cf. \cref{deriv-increasing-decreasing}); in partcular, it is injective. Moreover, we have
	\[ \lim_{x \to -1^+} f(x) = -\infty \quad \text{ and } \lim_{x \to 1^-} f(x) = +\infty \]
	so the intermediate value theorem guarantees that $f$ is bijective. Thus the inverse function $f^{-1} : \R \to (-1,1)$ exists. Finally, since $f'$ is always nonzero, it follows from \cref{smooth-nonzero-deriv-diffeo-single-var} that $f^{-1}$ must be smooth. It is possible to verify that
	\[ f^{-1}(x) = \frac{-1+\sqrt{4x^2+1}}{2x} \]
	by check that composing this formula with $f$ yields the identity, but, thanks to \cref{smooth-nonzero-deriv-diffeo-single-var}, we don't actually need this formula for $f^{-1}$ at all to know that $f^{-1}$ is smooth.
\end{example}

\begin{lemma} \label{smooth-nonzero-deriv-diffeo-single-var}
	Suppose $U$ and $V$ are open subsets of $\R$, that $f : U \to V$ is a smooth bijection, and that $f'$ is always nonzero. Then $f$ is a diffeomorphism. 
\end{lemma}

\begin{proof}
	We want to show that $f^{-1} : V \to U$ is smooth. Since $f'$ is always nonzero, we know from \cref{inverse-derivative} that $f^{-1}$ is differentiable and that
	\[ (f^{-1})'(a) = \frac{1}{f'(f^{-1}(a))}. \]
	In other words, $(f^{-1})'$ is the composite of three functions: $f^{-1}$, $f'$, and inversion (ie, the function $x \mapsto 1/x$). Inversion is clearly smooth; since $f$ is smooth, so is $f'$. 
	
	We now prove by induction that $f^{-1}$ is $C^k$ for all integers $k$. For the base case, we know from above that it is $C^0$ (in fact, we even know more: we know that $f^{-1}$ is even differentiable, but all we need to know for the base case is that it is continuous). Suppose inductively that $f^{-1}$ is $C^k$ for some non-negative integer $k$. Then, as we saw above, $(f^{-1})'$ is the composite of $f^{-1}$, $f'$, and inversion. The latter two are smooth, so in particular are $C^k$. By the inductive hypothesis, we know that $f^{-1}$ is $C^k$. Thus $(f^{-1})'$, being the composite of three $C^k$ functions, is also $C^k$ by \cref{composite-cn-singlevar}. This means that $f^{-1}$ is $C^{k+1}$, which completes the induction. 
\end{proof}

\begin{exercise}
	For any $a < b$, construct a diffeomorphism $f : (a, b) \to \R$.
\end{exercise}

Our next example of a slightly strange smooth function is the following one, which gets ``infinitely flat'' at 0 but is not constantly equal to 0. 

\begin{remark} \label{exponential-limit}
	Here is a fact that you might recognize:
	\begin{equation} \label{exponential-polynomial} \lim_{u \to \infty} \frac{u^m}{e^u} = 0 \end{equation}
	for any non-negative integer $m$. We'll use this fact, and also the related fact that
	\[ \lim_{x \to 0^+} \frac{e^{-1/x}}{x^m} = 0 \]
	for any non-negative integer $m$, which follows immediately from \cref{exponential-polynomial} by making the substitution $u = 1/x$.
	
	In your calculus class, you probably remember proving \cref{exponential-polynomial} using L'H\^{o}pital's rule. We haven't proved  L'H\^{o}pital's rule (except for the weak version in \cref{weak-lhopital}, which is only enough to calculate \cref{exponential-polynomial} for $m = 1$), but you can find proofs of L'H\^opital's rule in many places (eg, \cite[theorem 4.15]{protter-morrey}, \cite[theorem 5.13]{rudin}, or even \href{https://en.wikipedia.org/wiki/L\%27hopital\%27s_rule}{Wikipedia}). The proof is a clever application of the mean value theorem \ref{mean-value-theorem}. 
	
	An alternative proof of \cref{exponential-polynomial} uses the power series definition of the exponential function. The idea is to notice that
	\[ \frac{e^u}{u^m} = \frac{1}{u^m} + \dotsb + \frac{1}{(m-1)!u} + \sum_{k = 0}^\infty \frac{u^k}{(m+k)!} \]
	As $u \to \infty$, the first few summands all tend to 0, and the series at the end tends to $\infty$. \Cref{exponential-polynomial} follows from this. Details are omitted, but you should be able to work them out yourself if you're interested.
\end{remark} 

\begin{exercise} \label{infinitely-flat}
	Consider the function $f : \R \to \R$ defined as follows.
	\[ f(x) = \begin{cases} e^{-1/x} & \text{if } x > 0 \\ 0 & \text{if } x \leq 0 \end{cases} \]
	See \cref{infinitely-flat-graph}. 
	\begin{figure}
		\begin{center}
			\begin{tikzpicture}
			\begin{axis}[
			xmin=-1,xmax=2,
			ymin=-0.25,ymax=1.1,
			ytick={1},
			xtick={-1,1},
			xticklabels={$-1$,1},
			yticklabels={1},
			axis lines=middle,
			axis line style=lightgray,
			width=0.75\textwidth
			]
			
			
			\addplot[black,style=thick,->] expression[domain=0.01:1.9,samples=100]{exp(-1/x)};
			\addplot[black,style=thick,<-] expression[domain=-0.9:0,samples=2]{0};
			\end{axis}
			\end{tikzpicture}
		\end{center}
		\caption{The graph of the function $f$ from \cref{infinitely-flat}.}  \label{infinitely-flat-graph}
	\end{figure}
	\begin{enumerate}[(a)]
		\item Use \cref{exponential-limit} to show, using the definition of the derivative \ref{single-var-derivative-definition}, that $f'(0) = 0$.
		\item Check that
		\[ f'(x) = \begin{cases} \dfrac{e^{-1/x}}{x^2} & \text{if } x > 0 \\ 0 & \text{if } x \leq 0. \end{cases} \]
		You've already computed $f'(0)$. You can compute formulas for the derivative $f'(x)$ for $x > 0$ and for $x < 0$ using the usual rules for differentiation.
		\item Use \cref{exponential-limit} to show that $f''(0) = 0$, and then check that 
		\[ f''(x) = \begin{cases} p_2(x) \cdot \dfrac{e^{-1/x}}{x^4} & \text{if } x > 0 \\ 0 & \text{if } x \leq 0 \end{cases} \]
		where $p_2$ is the function $p_2(x) = 1-2x$. 
		\item Inductively, prove that
		\[ f^{(k)}(x) = \begin{cases} p_k(x) \cdot \dfrac{e^{-1/x}}{x^{2k}} & \text{if } x > 0 \\ 0 & \text{if } x \leq 0 \end{cases} \]
		where $p_k$ is a polynomial function of degree $k-1$ such that $p_k(0) = 1$. 
	\end{enumerate}
	Thus $f$ is an example of a smooth function that is ``infinitely flat'' at $x = 0$, in the sense that $f^{(k)}(x) = 0$ for all $k$, but is not constantly equal to 0. 
\end{exercise}

Next up, we have ``bump functions.'' Recall that the ``support'' of a function $f : U \to \R$ is the closure of the set $\{ x \in U : f(x) \neq 0 \}$. A ``bump function'' is a smooth function whose support is compact. 

\begin{exercise} \label{canonical-bump}
	Consider the function $\psi : \R \to \R$ given by the following. 
	\[ \psi(x) = \begin{cases} e^{-1/(1-x^2)} & \text{if } x \in (-1,1) \\ 0 & \text{otherwise}. \end{cases} \]
	See \cref{canonical-bump-graph} for its graph. 
	\begin{figure}
		\begin{center}
			\begin{tikzpicture}
			\begin{axis}[
			xmin=-1.5,xmax=1.5,
			ymin=-0.1,ymax=0.5,
			ytick={0},
			xtick={-1,1},
			xticklabels={$-1$,1},
			yticklabels={},
			axis lines=middle,
			axis line style=lightgray,
			width=0.75\textwidth
			]
			
			
			\addplot[black,style=thick] expression[domain=-1:1,samples=100]{exp(-1/(1-x^2))}; 
			
			\addplot[black,style=thick,<-] expression[domain=-1.5:-1,samples=2]{0};
			
			\addplot[black,style=thick,->] expression[domain=1:1.5,samples=2]{0};
			\end{axis}
			\end{tikzpicture}
		\end{center}
		\caption{The graph of the function $\psi$ from \cref{canonical-bump}.}  \label{canonical-bump-graph}
	\end{figure}
	Check that $\psi$ is the composite of the ``infinitely flat'' function from \cref{infinitely-flat} and the polynomial $p(x) = 1-x^2$. Conclude that $\psi$ is a bump function whose support is $[-1,1]$.
\end{exercise}

\begin{exercise}
	For any $a < b$ in $\R$, construct a bump function whose support is $[a,b]$. 
\end{exercise}

\begin{exercise}
	Let $F$ be a closed subset of $\R$. Construct a smooth function which is zero on $F$ and nonzero outside of $F$. \textit{Hint}. Recall that the complement of $F$ is a countable union of disjoint open intervals (cf. \cite[theorem 6.17]{protter-morrey}). 
\end{exercise}

\begin{comment}
The next example to consider is the so-called Gaussian function. 

\begin{example}[Gaussian function] \label{gaussian}
	Consider the function $\psi : \R \to \R$ defined by \[ \psi(x) = e^{-x^2}, \]
	called the \emph{Gaussian function}. As the composite of the smooth function $x \mapsto e^x$ with the smooth function $x \mapsto -x^2$, this function is also smooth. Observe that $\psi$ has the property that
	\[ \lim_{x \to \pm \infty} \psi(x) = 0. \]
	We can also calculate that $\psi'(x) = -2xe^{-x^2}$, and then using l'H\^opital's rule (which we haven't proved), we can show that
	\[ \lim_{x \to \pm \infty} \psi'(x) = 0. \]
	A more difficult property of $\psi$ is that
	\[ \int_{-\infty}^\infty \psi(t)\, dt = \sqrt{\pi}. \]
\end{example}
\end{comment}

In the following two examples, we will have to invoke the fundamental theorem of calculus, even though we have not proved it (or even defined integrals, for that matter). If you're unhappy with using things you haven't proved, you can find rigorous discussions of integration in many places (eg, \cite[chapter 5]{protter-morrey}, \cite[chapter 6]{rudin}, etc).

\begin{example} \label{integral-canonical-bump}
	Let $\psi$ be the bump function from \cref{canonical-bump}, and then consider 
	\[ \eta(x) = \int_{-1}^x \psi(t)\, dt. \]
	Then 
	\[ \eta'(x) = \frac{d}{dx} \int_{-1}^x \psi(t)\, dt = \psi(x) \]
	by the fundamental theorem of calculus. Thus $\eta$ is smooth, since $\eta' = \psi$ is smooth. Note moreover that $\eta$ is increasing, since $\eta' = \psi \geq 0$. Finally, it is clear from the geometric interpretation of integrals as ``area under the curve'' that $\eta$ is constantly 0 for all $x \leq -1$ and that it is constantly $\eta(1)$ for all $x \geq 1$. 
\end{example}

\begin{exercise} \label{constant-compact-zero-outside-neighborhood}
	Let $I = [a, b]$ be a compact interval in $\R$ and let $U$ be an open subset containing $I$. Prove that there exists a bump function $f : \R \to \R$ such that $f(x) = 1$ for all $x \in K$ and $f(x) = 0$ for all $x \notin U$. \textit{Hint.} The characteristic function \[ \chi_I(x) = \begin{cases} 1 & \text{if } x \in I \\ 0 & \text{if } x \notin I \end{cases} \]
	is close to having all of the properties we want, but it's not smooth (it's clearly not even continuous). Since $U$ is open, there exist $a' < a$ and $b' > b$ such that $I := [a', b'] \subseteq U$. Modify $\chi_I$ on the intervals $[a', a]$ and $[b, b']$ to obtain the desired function. Use inspiration from the function $\eta$ of \cref{integral-canonical-bump}.
\end{exercise}